---
title: "RepDB Comparison with NR"
date: "`r format(Sys.time(), '%d-%b-%Y')`"
author: "Giacomo Mutti"
output:
    rmdformats::html_clean:
        number_sections: false
        code_folding: hide
        self_contained: true
        fig_caption: true
        # thumbnails: false
        # lightbox: false
        gallery: true
        toc_depth: 3
        highlight: kate
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, warning = FALSE)
knitr::knit_hooks$set(crop = knitr::hook_pdfcrop)
knitr::opts_chunk$set(crop = TRUE)
```

```{r libs}
library(tidyverse)
library(tidyheatmaps)
library(patchwork)
library(ggalluvial)

theme_set(theme_classic())
source("workflow/scripts/functions.R")
source("workflow/scripts/palettes.R")
```
```{r read_stuff}
df_dmnd <- read_delim(list.files("results/comparison/hits/", pattern = "matches.tsv", full.names = T),
  id = "db",
  col_names = dmnd_cols
) %>%
  mutate(db = gsub("_matches.tsv", "", basename(db))) %>%
  mutate(db = factor(db, levels = names(color_benchdbs)))


# repdb_tax <- read_delim("results/taxonomies/repdb.tsv",
#            col_names = c("mnemo", "k", "p", "c", "o", "f", "g", "s"))
# min_cov <- 50
# map <- read_delim("test/blast_search/tax.tsv",
#                   col_names = c("sseqid", "taxid", "k", "p", "c", "o", "f", "g", "s")) %>%
#   distinct()

# df_dmnd <- df %>%
# left_join(map) %>%
# mutate(mnemo = str_split_i(sseqid, "_", 2)) %>%
# left_join(repdb_tax, by="mnemo") %>%
# mutate(
#   k = coalesce(k.x, k.y),
#   p = coalesce(p.x, p.y),
#   c = coalesce(c.x, c.y),
#   o = coalesce(o.x, o.y),
#   f = coalesce(f.x, f.y),
#   g = coalesce(g.x, g.y),
#   s = coalesce(s.x, s.y)
# ) %>%
# select(-ends_with(".x"), -ends_with(".y")) %>%
# filter(qcovhsp >= min_cov)
```

```{r benchmark}
bench_df <- list.files("results/benchmarks/comparison/", full.names = T) %>%
  read_delim(id = "tmp") %>%
  separate(tmp, c("db", "method"), "_") %>%
  mutate(
    db = basename(db),
    method = gsub(".txt", "", method),
    max_rss_gb = max_rss / 1024,
    db_source = gsub("clust", "", db)
  )

bench_df %>%
  filter(!grepl("clust", db)) %>%
  select(db, method, s, max_rss_gb) %>%
  pivot_wider(names_from = db, values_from = c(s, max_rss_gb)) %>%
  mutate(
    s_ratio = s_nr / s_repdb,
    max_rss_gb_ratio = max_rss_gb_nr / max_rss_gb_repdb
  )

plot_bench <- bench_df %>%
  ggplot(aes(max_rss_gb, `h:m:s`, fill = db, group = db_source)) +
  geom_line(color = "grey60") +
  geom_point(pch = 21, size = 2) +
  facet_grid(~method, scales = "free") +
  labs(x = "Mem (GB)", y = "Runtime") +
  scale_fill_manual(values = color_benchdbs)
```

```{r diamond_results}
plot_dmnd <- df_dmnd %>%
  select(db, pident, qcovhsp) %>%
  mutate(
    clustered = grepl("clust", db),
    db2 = gsub("clust", "", db)
  ) %>%
  pivot_longer(cols = c(pident, qcovhsp)) %>%
  ggplot(aes(value, color = db)) +
  geom_density() +
  facet_grid(. ~ name, scales = "free_y") +
  labs(x = "") +
  scale_color_manual(values = color_benchdbs) +
  theme(legend.position = "none")
```

```{r mmseqs}
order_ranks <- c(
  "no rank", "cellular root", "domain", "kingdom", "phylum", "division",
  "class", "clade", "order", "family", "genus", "species", "strain"
)

df_mmseqs_tmp <- read_delim(
  list.files("results/comparison/tax/",
    pattern = "lca.tsv",
    full.names = T
  ),
  col_names = c("seq", "taxid", "rank", "classification"),
  id = "db"
) %>%
  mutate(db = gsub("_lca.tsv", "", basename(db)))

df_cohen <- read_delim("resources/Cohen_2024/merged.fasta.transdecoder-estimated-taxonomy.LCA.phyloDB.within3percent.AddRef.RefSeq.out") %>%
  filter(transcript_name %in% df_mmseqs_tmp$seq)

df_mmseqs <- df_cohen %>%
  select(transcript_name, classification_level, classification) %>%
  mutate(db = "Cohen2024", taxid = NA) %>%
  rename(
    "seq" = "transcript_name",
    "rank" = "classification_level",
    # "classification" = "full_classification"
  ) %>%
  bind_rows(df_mmseqs_tmp) %>%
  mutate(rank = case_when(db == "Cohen2024" & rank == "supergroup" ~ "domain",
    grepl("repdb", db) & rank == "superkingdom" ~ "domain",
    .default = rank
  )) %>%
  mutate(
    db = factor(db, levels = c("Cohen2024", names(color_benchdbs))),
    rank = factor(rank, levels = rev(order_ranks))
  )
```

```{r mmseqs-kraken}
kraken_files <- list.files("results/comparison/tax", full.names = TRUE, pattern = "_report")
kraken_files <- kraken_files[!str_detect(kraken_files, "tophit")]
krk <- read_delim(kraken_files,
  id = "db",
  col_names = c(
    "perc", "n", "n_direct", "rank",
    "taxid", "nm"
  ), delim = "\t"
) %>%
  mutate(
    nested = str_count(nm, "\\G ") / 2,
    nm = trimws(nm), nm_g = gsub(" .*", "", nm),
    db = gsub("_report", "", basename(db))
  )
```

```{r plot_kingdoms}
cohen_cnts <- df_cohen %>%
  mutate(
    db = "Cohen2024",
    nm = gsub(";..*", "", full_classification),
    nm = replace_na(nm, "unclassified")
  ) %>%
  count(db, nm) %>%
  mutate(nm = factor(nm, levels = c("unclassified", names(color_kingdoms)))) %>%
  arrange(nm) %>%
  mutate(y = n / cumsum(n))


plot_kings <- krk %>%
  filter(
    (grepl("nr", db) & nested < 3 & nm != "root") |
      (grepl("rep", db) & nested < 2 & nm != "root"),
    rank != "cellular root"
  ) %>%
  select(db, n, nm) %>%
  bind_rows(cohen_cnts) %>%
  mutate(
    nm = case_when(nm == "Uroviricota" ~ "Viruses", .default = nm),
    db = factor(db, levels = c(names(color_benchdbs), "Cohen2024")),
    nm = factor(nm, levels = c("unclassified", names(color_kingdoms)))
  ) %>%
  ggplot(aes(db, n, fill = nm)) +
  geom_bar(position = "fill", stat = "identity") +
  # geom_text(aes(label = nm), position = position_fill(vjust = .5), data = cohen_cnts) +
  labs(x = "", y = "Prop.", fill = "Kingdom") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.01))) +
  scale_fill_manual(values = c("unclassified" = "grey40", color_kingdoms)) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
    # legend.position = "none"
  )
```
```{r fullplot, fig.width=6.5, fig.height=4.5}
plot_comp <- ((plot_bench / plot_dmnd) | plot_kings) +
  plot_layout(widths = c(2, 1), guides = "collect")

saveRDS(plot_comp, "results/plots/comparison.Rds")
# ggsave("results/plots/comparison.pdf", plot_comp, width = 6.5, height = 4.5)
```


```{r alluvial}
df_classification <- df_mmseqs %>%
  filter(!grepl("clust", db)) %>%
  mutate(classification = replace_na(classification, "unclassified")) %>%
  select(seq, classification, db) %>%
  # filter(db=="Cohen2024") %>% count(classification) %>% View
  pivot_wider(names_from = db, values_from = classification)

df_classification %>% 
  filter(repdb=="unclassified")

p_alluvial <- df_classification %>%
  # filter(if_any(everything(), ~ .x == "unclassified")) %>%
  filter(repdb == "unclassified" | nr == "unclassified") %>%
  # filter(grepl("Dinophy", Cohen2024)) %>%
  ggplot(aes(axis1 = nr, axis2 = Cohen2024, axis3 = repdb)) +
  geom_alluvium(
    # aes(fill=Cohen2024),
    # color="grey70",
    fill = "grey80"
  ) +
  geom_stratum(
    aes(fill = after_stat(stratum) == "unclassified")
    # fill="grey"
  ) +
  geom_text(
    stat = "stratum",
    aes(label = after_stat(stratum)),
    color = "black",
    size = 1
  ) +
  scale_x_discrete(limits = c("nr", "Cohen2024", "repdb"), expand = c(.1, .1)) +
  labs(y = "N") +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("grey", "black"))

p_alluvial
# ggsave("results/plots/alluvial.pdf", p_alluvial, width = 5, height = 7)
```



## What is in the gap? (taxonomic composition of RepDB misses)
 
Because the gap set is defined on transcripts the *reference* classified, we can
describe it entirely in the reference's own labels — no harmonization needed. If
the gaps pile up in a clade RepDB filtered out (comment #2), it shows here.
 
```{r gap-composition}
cohen_lineage <- df_cohen %>%
  mutate(
    seq = transcript_name,
    domain = str_split_i(full_classification, ";\\s*", 1),
    supergroup = str_split_i(full_classification, ";\\s*", 2),
    division = str_split_i(full_classification, ";\\s*", 3)
  )



df_classification %>% 
  filter(Cohen2024!="unclassified") %>% 
  mutate(repdb = repdb!="unclassified") %>% 
  left_join(cohen_lineage, by = "seq") %>% 
  mutate(supergroup = replace_na(supergroup, "unassigned")) %>%
  count(repdb, domain, supergroup) %>%
  ggplot(aes(n, fct_reorder(supergroup, n), fill = repdb)) +
  geom_col(position = "stack") +
  facet_grid(domain ~ ., scales = "free_y", space = "free_y") +
  scale_x_continuous(expand = expansion(c(0,0.02))) + 
  scale_fill_manual(values = c("grey40" , color_benchdbs[["repdb"]])) +
  labs(x = "N transcripts", y = "", fill = "Classified in RepDB") +
  theme(legend.position = "bottom", strip.text = element_text(size=4))


```
 
## Why is a transcript unclassified? (no homolog vs no LCA)
 
Splitting RepDB's "unclassified" into *no DIAMOND hit at all* vs *hit but the LCA
did not assign a taxon* separates a genuine coverage gap (missing sequences, ties
to comment #2) from a framework/LCA artifact. Note the two engines use different
thresholds, so DIAMOND presence is a homology *proxy*.
 
```{r unclassified-reason}
hit_seqs <- df_dmnd %>%
  filter(!grepl("clust", db)) %>%
  distinct(db = as.character(db), qseqid) %>%
  mutate(has_hit = TRUE)

df_dmnd %>% 
  group_by(qseqid, db) %>% 
  summarise(n=n(), mn_ident = mean(pident)) %>% 
  ungroup() %>% 
  full_join(filter(df_mmseqs, db!="Cohen2024"), 
            by = c("qseqid"="seq", "db")) %>% 
  mutate(unclassified = classification=="unclassified", 
         no_hit = is.na(n)) %>% 
  count(db, unclassified, no_hit) %>% 
  filter(unclassified) %>% 
  ggplot(aes(db, n, fill=no_hit)) + 
  geom_col()
```
 
## Resolution: how deep does each DB classify?
 
Coverage isn't everything — a DB that classifies everything to "domain" isn't
useful. This is the specificity side of the trade-off.
 
```{r resolution-profile}
(plot_res <- df_mmseqs %>%
  filter(!grepl("^clust", db), !is.na(rank)) %>%
  count(db, rank) %>%
  group_by(db) %>% mutate(prop = n / sum(n)) %>% ungroup() %>%
  ggplot(aes(prop, rank, fill = db)) +
  geom_col(position = "dodge") +
   facet_wrap(~db, scales = "free_y") +
  scale_fill_manual(values = c("Cohen2024" = "firebrick", color_benchdbs)) +
  labs(x = "Proportion of classified transcripts", 
       y = "Assigned rank", fill = "")) +
  theme(legend.position = "none")
```
 
## broad-agreement
 
```{r agreement}
df_classification %>% 
  count(repdb, Cohen2024) %>% 
  filter(n>2) %>% 
  ggplot(aes(repdb, Cohen2024, fill=n))  + 
  geom_tile() + 
  scale_fill_viridis_c() + 
  theme(axis.text.x = element_text(angle=60, hjust=1))

df_classification %>% 
  count(repdb, nr) %>% 
  filter(n>2) %>% 
  ggplot(aes(repdb, nr, fill=n))  + 
  geom_tile() + 
  scale_fill_viridis_c() + 
  theme(axis.text.x = element_text(angle=60, hjust=1))

df_classification %>% 
  count(Cohen2024, nr) %>% 
  filter(n>1) %>% 
  ggplot(aes(Cohen2024, nr, fill=n))  + 
  geom_tile() + 
  scale_fill_viridis_c() + 
  theme(axis.text.x = element_text(angle=60, hjust=1))
```

# TODOs
 
* Analyze taxonomic distribution of DIAMOND results.
* Analyze taxonomic distribution of diamond results?
* Compare mmseqs easy-taxonomy results
