---
title: "RepDB vs nr: search cost, coverage and taxonomic resolution"
date: "`r format(Sys.time(), '%d-%b-%Y')`"
author: "Giacomo Mutti"
output:
    rmdformats::html_clean:
        number_sections: false
        code_folding: hide
        self_contained: true
        fig_caption: true
        gallery: true
        toc_depth: 3
        highlight: kate
---

<!--
Render from the repository root:

  Rscript -e 'root <- getwd()
  dir.create(file.path(root, "results/qc"), recursive = TRUE, showWarnings = FALSE)
  rmarkdown::render("workflow/notebooks/comparison.Rmd",
                    knit_root_dir = root,
                    output_dir    = file.path(root, "results/qc"),
                    output_file   = "comparison.html")'

(Bind the root to a variable first: render() setwd()s into the notebook's
directory before it forces `output_file`.)
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, warning = FALSE, message = FALSE)
if (nzchar(Sys.which("pdfcrop"))) {
  knitr::knit_hooks$set(crop = knitr::hook_pdfcrop)
  knitr::opts_chunk$set(crop = TRUE)
}
```

```{r libs}
suppressPackageStartupMessages({
  library(tidyverse)
  library(patchwork)
})
# free() is used to lay out Fig. 5 and only exists from patchwork 1.2.0
stopifnot(packageVersion("patchwork") >= "1.2.0")
theme_set(theme_classic(base_size = 9))

.src <- function(p) source(if (file.exists(p)) p else file.path("..", "..", p))
.src("workflow/scripts/functions.R")
.src("workflow/scripts/palettes.R")
```

```{r config}
P <- list(
  results   = "results",
  resources = "resources",
  figdir    = "results/plots",
  reference      = "resources/Cohen_2024/merged.fasta.transdecoder_reduced.tsv",
  reference_name = "Cohen2024",
  reference_long = "Cohen et al. (2024)"
)
```

```{r paths}
res <- function(...) rp(file.path(P$results, ...))
rsc <- function(...) rp(file.path(P$resources, ...))
have <- function(p) !is.null(p) && length(p) == 1L && nzchar(p) && file.exists(rp(p))

figdir <- rp(P$figdir)
dir.create(figdir, showWarnings = FALSE, recursive = TRUE)
fig <- function(name) file.path(figdir, name)

REF <- P$reference_name       # short, for axes and legends
REF_LONG <- P$reference_long  # full, for prose

color_ref <- c("#CD5D8E")
names(color_ref) <- REF
db_colors <- c(color_benchdbs, color_ref)
DB_LEVELS <- names(db_colors)

scale_db <- function() {
  scale_color_manual(values = db_colors, limits = DB_LEVELS, drop = FALSE,
                     na.translate = FALSE,
                     guide = guide_legend(direction = "vertical"))
}

OUTCOMES <- c(sprintf("Agrees with %s", REF), "Assigned, differs",
              "Homolog, no LCA", "No homolog")
outcome_colors <- setNames(outcome_colors_ref, OUTCOMES)

N <- list()

thm <- theme_classic(base_size = 10) +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(size = 10),
    legend.key.size = unit(4.5, "mm"),
    legend.text = element_text(size = 9),
    legend.title = element_blank(),
    axis.text = element_text(size = 9),
    plot.tag = element_text(size = 12, face = "bold"),
    # panels A and B are facetted on free scales, so both facets run their axis
    # out to a wide tick; at 9pt the last label of the left facet and the first
    # of the right one collide unless the strips are held apart
    panel.spacing = unit(4, "mm"),
    plot.margin = margin(3, 3, 3, 3)
  )

show_tbl <- function(df, caption = NULL, digits = 1) {
  k <- knitr::kable(df, caption = caption, digits = digits,
                    format.args = list(big.mark = ","))
  if (requireNamespace("kableExtra", quietly = TRUE)) {
    k <- kableExtra::kable_styling(k, bootstrap_options = c("striped", "condensed"),
                                   full_width = FALSE)
  }
  k
}
fmt_n <- function(x) formatC(round(x), format = "d", big.mark = ",")
fmt_p <- function(x, d = 1) paste0(formatC(x, format = "f", digits = d), "%")
fmt_d <- function(x, d = 2) formatC(x, format = "f", digits = d)
fmt_x <- function(x, d = 2) paste0(formatC(x, format = "f", digits = d), "x")
```

# Input validation

Every per-transcript result must refer to the *current* query set. This gate
exists because `mmseqs easy-taxonomy` writes four files per run and it is easy to
end up with a `_lca.tsv` left over from an earlier sample while the `_report`
next to it is current — the two disagree silently and every downstream
"unclassified" number is then wrong.

```{r validate}
fa <- readLines(res("comparison", "input.fa"), warn = FALSE)
query_ids <- sub(" .*", "", sub("^>", "", fa[startsWith(fa, ">")]))
N$n_queries <- length(query_ids)

# per-transcript LCA assignments
lca_files <- list.files(res("comparison", "tax"), pattern = "_lca\\.tsv$", full.names = TRUE)
lca_all <- map_dfr(lca_files, ~ read_delim(.x,
  col_names = c("seq", "taxid", "rank", "classification"),
  show_col_types = FALSE, progress = FALSE
) %>% mutate(db = gsub("_lca.tsv", "", basename(.x))))

# kraken-style aggregate reports (one row per taxon, depth-indented)
report_files <- list.files(res("comparison", "tax"), pattern = "_report$", full.names = TRUE)
report_files <- report_files[!str_detect(report_files, "tophit")]
reports <- map_dfr(report_files, ~ read_delim(.x, delim = "\t",
  col_names = c("perc", "n", "n_direct", "rank", "taxid", "nm"),
  show_col_types = FALSE, progress = FALSE
) %>% mutate(db = gsub("_report", "", basename(.x))))

validation <- bind_rows(
  lca_all %>% group_by(db) %>%
    summarise(File = "_lca.tsv", Rows = n(),
              `Matching current input` = sum(seq %in% query_ids), .groups = "drop"),
  reports %>% group_by(db) %>%
    summarise(File = "_report", Rows = sum(n_direct),
              `Matching current input` = NA_integer_, .groups = "drop")
) %>%
  mutate(
    `% current` = `Matching current input` / Rows * 100,
    Status = case_when(
      File == "_report" & Rows == N$n_queries ~ "OK",
      File == "_report" ~ "STALE (row total != query count)",
      `% current` > 99 ~ "OK",
      TRUE ~ "STALE - re-run mmseqs for this db"
    )
  ) %>%
  select(db, File, Rows, `Matching current input`, `% current`, Status) %>%
  arrange(File, db)

# databases whose per-transcript file can be trusted
LCA_DBS <- lca_all %>%
  group_by(db) %>%
  summarise(ok = mean(seq %in% query_ids) > 0.99) %>%
  filter(ok) %>% pull(db)
REPORT_DBS <- reports %>%
  group_by(db) %>% summarise(ok = sum(n_direct) == N$n_queries) %>%
  filter(ok) %>% pull(db)

lca <- filter(lca_all, db %in% LCA_DBS)
FULL_LCA <- setequal(LCA_DBS, unique(lca_all$db))

show_tbl(validation, sprintf("Consistency of each result file with the %s queries in `input.fa`.", fmt_n(N$n_queries)), digits = 1)
```

```{r validate-note, echo=FALSE, results="asis"}
if (!FULL_LCA) {
  cat(sprintf(
    "> **Warning**: the per-transcript `_lca.tsv` is stale for **%s**. The aggregate `_report` files are current for all databases, so the figure panels, the coverage numbers and the resolution profile below use *all four* databases. The analyses that need a per-transcript assignment (agreement with the reference, the gap decomposition) are restricted to **%s**. Re-run `mmseqs easy-taxonomy` (or re-copy the outputs) for the stale databases to complete them.\n\n",
    paste(setdiff(unique(lca_all$db), LCA_DBS), collapse = ", "),
    paste(LCA_DBS, collapse = ", ")
  ))
}
```

```{r read-rest}
# DIAMOND hits: only the columns needed, the files are ~150-190 MB each
hits <- map_dfr(
  list.files(res("comparison", "hits"), pattern = "matches.tsv", full.names = TRUE),
  function(p) {
    read_delim(p, col_names = FALSE, col_select = c(1, 3, 13),
               show_col_types = FALSE, progress = FALSE) %>%
      set_names(c("qseqid", "pident", "qcovhsp")) %>%
      mutate(db = gsub("_matches.tsv", "", basename(p)))
  }
) %>% mutate(db = factor(db, levels = DB_LEVELS))

# reference annotation of the same transcripts
ref <- read_delim(rp(P$reference), show_col_types = FALSE, progress = FALSE) %>%
  transmute(
    seq = transcript_name,
    ref_rank = classification_level,
    ref_taxon = classification,
    ref_lineage = full_classification,
    ambiguous
  ) %>%
  distinct(seq, .keep_all = TRUE) %>%
  mutate(
    ref_domain     = str_split_i(ref_lineage, ";\\s*", 1),
    ref_supergroup = str_split_i(ref_lineage, ";\\s*", 2),
    ref_division   = str_split_i(ref_lineage, ";\\s*", 3)
  )

N$n_ref_total <- nrow(ref)
N$n_ref_unclassified <- sum(is.na(ref$ref_taxon) | ref$ref_taxon == "unclassified")
N$n_ref_ambiguous <- sum(ref$ambiguous == 1, na.rm = TRUE)
```

# Fig. 5

## A - Search cost

```{r benchmarks}
bench <- list.files(res("benchmarks", "comparison"), full.names = TRUE, pattern = "\\.txt$") %>%
  map_dfr(~ read_delim(.x, delim = "\t", show_col_types = FALSE, progress = FALSE) %>%
    mutate(tag = str_remove(basename(.x), "\\.txt$"))) %>%
  separate(tag, c("db", "tool"), sep = "_(?=[^_]+$)") %>%
  mutate(
    tool = recode(tool, dmnd = "DIAMOND", mmseqs = "MMseqs2"),
    mem_gb = max_rss / 1024,
    minutes = s / 60,
    db_source = gsub("^clust", "", db),
    clustered = str_starts(db, "clust"),
    db = factor(db, levels = DB_LEVELS)
  )

cost <- bench %>%
  select(tool, db, minutes, mem_gb) %>%
  arrange(tool, db)

show_tbl(
  cost %>% rename(Tool = tool, DB = db, `Runtime (min)` = minutes, `Peak RSS (GB)` = mem_gb),
  "Wall time and peak memory of the homology search / taxonomic assignment on the full query set."
)

ratio <- cost %>%
  filter(db %in% c("nr", "repdb")) %>%
  pivot_wider(names_from = db, values_from = c(minutes, mem_gb)) %>%
  mutate(`Runtime speedup` = minutes_nr / minutes_repdb,
         `Memory reduction` = mem_gb_nr / mem_gb_repdb)
ratio_c <- cost %>%
  filter(db %in% c("clustnr", "clustrepdb")) %>%
  pivot_wider(names_from = db, values_from = c(minutes, mem_gb)) %>%
  mutate(`Runtime speedup` = minutes_clustnr / minutes_clustrepdb,
         `Memory reduction` = mem_gb_clustnr / mem_gb_clustrepdb)

gv <- function(df, tool, col) df[[col]][df$tool == tool]
N$speed_dmnd  <- gv(ratio, "DIAMOND", "Runtime speedup")
N$speed_mmseqs <- gv(ratio, "MMseqs2", "Runtime speedup")
N$mem_dmnd    <- gv(ratio, "DIAMOND", "Memory reduction")
N$mem_mmseqs  <- gv(ratio, "MMseqs2", "Memory reduction")
N$mem_mmseqs_nr    <- gv(filter(cost, db == "nr"), "MMseqs2", "mem_gb")
N$mem_mmseqs_repdb <- gv(filter(cost, db == "repdb"), "MMseqs2", "mem_gb")

show_tbl(
  bind_rows(
    ratio %>% transmute(Comparison = "RepDB vs nr", Tool = tool, `Runtime speedup`, `Memory reduction`),
    ratio_c %>% transmute(Comparison = "ClusteredRepDB vs ClusteredNR", Tool = tool, `Runtime speedup`, `Memory reduction`)
  ),
  "Ratios (>1 means RepDB is faster / leaner).", digits = 2
)

# A and B suppress their keys and D carries the single collected legend, so the
# fill scale here never has to merge with D's colour scale
plot_bench <- bench %>%
  ggplot(aes(mem_gb, minutes, fill = db, group = db_source)) +
  geom_line(color = "grey75", linewidth = .3) +
  geom_point(size = 2.5, pch=21) +
  facet_wrap(~tool, scales = "free") +
  scale_fill_manual(values = db_colors) +
  # the memory axes span very different ranges; give the outermost tick room so
  # it is not clipped by the panel edge in the narrow left column
  scale_x_continuous(n.breaks = 4, expand = expansion(mult = 0.12)) +
  labs(x = "Peak memory (GB)", y = "Runtime (min)") +
  thm + theme(legend.position = "none")   # the shared key lives on panel D
```

Against nr, RepDB is **`r fmt_x(N$speed_dmnd)` faster for DIAMOND** and
**`r fmt_x(N$speed_mmseqs)` faster for MMseqs2**. DIAMOND memory drops
`r fmt_x(N$mem_dmnd)`, but MMseqs2 peak memory is essentially unchanged
(`r fmt_d(N$mem_mmseqs_repdb, 0)` GB vs `r fmt_d(N$mem_mmseqs_nr, 0)` GB): the
`easy-taxonomy` prefilter sizes its working set to the available RAM rather than
to the database, so database size does not translate into a memory saving there.
Clustering is what reduces it (see the table above).

## B - Homology search results

```{r diamond}
best <- hits %>%
  group_by(db, qseqid) %>%
  slice_max(pident, n = 1, with_ties = FALSE) %>%
  ungroup()

coverage <- hits %>%
  group_by(db) %>%
  summarise(
    Hits = n(),
    `Queries with a hit` = n_distinct(qseqid),
    `% of queries` = n_distinct(qseqid) / N$n_queries * 100,
    `Hits per query` = n() / n_distinct(qseqid),
    `Median identity` = median(pident),
    `Median coverage` = median(qcovhsp),
    .groups = "drop"
  )

N$hit_repdb <- coverage$`% of queries`[coverage$db == "repdb"]
N$hit_nr    <- coverage$`% of queries`[coverage$db == "nr"]
N$nohit_repdb <- N$n_queries - coverage$`Queries with a hit`[coverage$db == "repdb"]

show_tbl(coverage, "DIAMOND homology search: how many of the queries each database can place at all.")

plot_dmnd <- best %>%
  select(db, pident, qcovhsp) %>%
  pivot_longer(c(pident, qcovhsp)) %>%
  mutate(name = factor(recode(name, pident = "% identity", qcovhsp = "% coverage"),
                       levels = c("% identity", "% coverage"))) %>%
  ggplot(aes(value, color = db)) +
  geom_density(linewidth = .5, key_glyph = "point") +
  facet_wrap(~name, scales = "free") +
  # both facets run to 100; without room at the edge that last tick label is
  # clipped by the panel border
  scale_x_continuous(expand = expansion(mult = 0.06)) +
  scale_db() +
  labs(x = "", y = "Density") +
  thm + theme(legend.position = "none")   # the shared key lives on panel D
```

**RepDB finds a homolog for `r fmt_p(N$hit_repdb)` of the queries against nr's
`r fmt_p(N$hit_nr)`**, despite being roughly a third of the size. Only
`r fmt_n(N$nohit_repdb)` queries have no RepDB homolog at all.

```{r identity-bands}
id_bands <- function(d) {
  d %>%
    mutate(band = cut(pident, c(0, 30, 50, 70, 90, 100),
                      labels = c("<30", "30-50", "50-70", "70-90", ">90"))) %>%
    count(db, band) %>%
    group_by(db) %>% mutate(pct = n / sum(n) * 100) %>% ungroup() %>%
    select(-n) %>% pivot_wider(names_from = db, values_from = pct) %>%
    select(band, any_of(DB_LEVELS))
}

# best hit per query - what Fig. 5B shows, and what matters for annotation
bands <- id_bands(best)
# all hits - where the effect of clustering on redundancy is visible
bands_all <- id_bands(hits)

pick_band <- function(tbl, db, b) sum(tbl[[db]][tbl$band %in% b])
N$lowid_repdb <- pick_band(bands, "repdb", c("<30", "30-50"))
N$lowid_nr    <- pick_band(bands, "nr",    c("<30", "30-50"))
N$midid_repdb <- pick_band(bands, "repdb", "70-90")
N$midid_nr    <- pick_band(bands, "nr",    "70-90")
for (d in DB_LEVELS[DB_LEVELS %in% names(bands_all)]) {
  N[[paste0("hi90_", d)]] <- pick_band(bands_all, d, ">90")
}

show_tbl(bands, "Identity of the **best hit per query** (% of queries with a hit). This is what Fig. 5B shows.")
show_tbl(bands_all, "Identity of **all hits** (% of hits). Clustering is visible here rather than in the best-hit view: a query's best hit is usually retained as a cluster representative, so removing near-identical redundancy barely moves the best-hit distribution but strips most of the >90% identity hits from the full output.")
```

RepDB's best hits are *closer*: `r fmt_p(N$lowid_repdb)` fall below 50% identity
versus `r fmt_p(N$lowid_nr)` for nr, and `r fmt_p(N$midid_repdb)` lie between 70%
and 90% versus `r fmt_p(N$midid_nr)`. Query coverage, by contrast, is
indistinguishable between the two.

Across **all** hits, clustering strips the high-identity redundancy as intended:
hits above 90% identity fall from `r fmt_p(N$hi90_nr)` to `r fmt_p(N$hi90_clustnr)`
of the total in ClusteredNR, and from `r fmt_p(N$hi90_repdb)` to
`r fmt_p(N$hi90_clustrepdb)` in ClusteredRepDB.

## C - What happened to every query

Panels B and D describe the queries a database *does* place. This one accounts
for all of them, so the coverage rate, the agreement rate and the reason a
transcript ends up unlabelled can be read off a single decomposition. It is
restricted to the queries `r REF` itself placed in a domain, so every category
shares one denominator.

```{r outcome}
# lineages rebuilt from the depth-indented report, so an LCA taxid resolves to a
# domain without needing the taxdump
lineage_of <- function(rep) {
  rep <- rep %>% mutate(depth = (nchar(nm) - nchar(str_trim(nm, "left"))) %/% 2,
                        nm = str_trim(nm))
  stack <- character(0); out <- character(nrow(rep))
  for (i in seq_len(nrow(rep))) {
    d <- rep$depth[i]
    if (length(stack) > d) stack <- stack[seq_len(d)]
    stack <- c(stack, rep$nm[i])
    out[i] <- paste(stack, collapse = ";")
  }
  tibble(taxid = rep$taxid, lineage = out)
}
lin <- reports %>% filter(db %in% REPORT_DBS) %>%
  group_split(db) %>%
  map_dfr(~ lineage_of(.x) %>% mutate(db = .x$db[1]))

DOMAINS <- c("Bacteria", "Archaea", "Eukaryota", "Viruses")

agreement <- lca %>%
  left_join(distinct(lin, db, taxid, lineage), by = c("db", "taxid")) %>%
  mutate(db_domain = map_chr(str_split(lineage, ";"),
    ~ { h <- intersect(.x, DOMAINS); if (length(h)) h[1] else NA_character_ })) %>%
  inner_join(ref, by = "seq") %>%
  filter(ref_domain %in% DOMAINS)

outcome <- agreement %>%
  left_join(hits %>% distinct(db, qseqid) %>%
              mutate(db = as.character(db), has_hit = TRUE),
            by = c("db", "seq" = "qseqid")) %>%
  mutate(
    has_hit = replace_na(has_hit, FALSE),
    outcome = case_when(
      !is.na(db_domain) & db_domain == ref_domain ~ OUTCOMES[1],
      !is.na(db_domain)                           ~ OUTCOMES[2],
      has_hit                                     ~ OUTCOMES[3],
      TRUE                                        ~ OUTCOMES[4]
    )
  )

outcome_tbl <- outcome %>%
  count(db, outcome) %>%
  group_by(db) %>% mutate(pct = n / sum(n) * 100) %>% ungroup()

show_tbl(
  outcome_tbl %>% select(-n) %>%
    pivot_wider(names_from = outcome, values_from = pct) %>%
    rename(DB = db) %>% select(DB, any_of(OUTCOMES)),
  sprintf("What happened to each of the %s queries %s placed in a domain (%% of that set).",
          fmt_n(n_distinct(outcome$seq)), REF)
)

plot_outcome <- outcome_tbl %>%
  mutate(db = factor(db, levels = DB_LEVELS),
         outcome = factor(outcome, levels = rev(OUTCOMES))) %>%
  ggplot(aes(db, pct, fill = outcome)) +
  geom_col(position = "stack", width = .75) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.01))) +
  scale_fill_manual(values = outcome_colors, breaks = OUTCOMES,
                    guide = guide_legend(direction = "vertical")) +
  labs(x = "", y = "% of queries") +
  thm +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
```

## Kingdom composition (supplementary)

```{r kingdoms}
ref_kingdom <- ref %>%
  mutate(nm = replace_na(ref_domain, "unclassified")) %>%
  count(nm) %>% mutate(db = REF)

plot_kings <- reports %>%
  filter(db %in% REPORT_DBS) %>%
  mutate(depth = (nchar(nm) - nchar(str_trim(nm, "left"))) %/% 2, nm = str_trim(nm)) %>%
  filter(nm %in% c("Bacteria", "Archaea", "Eukaryota", "Viruses", "unclassified")) %>%
  group_by(db, nm) %>% summarise(n = sum(n), .groups = "drop") %>%
  bind_rows(ref_kingdom) %>%
  mutate(
    db = factor(db, levels = DB_LEVELS),
    nm = factor(nm, levels = c("unclassified", names(color_kingdoms)))
  ) %>%
  ggplot(aes(db, n, fill = nm)) +
  geom_col(position = "fill", width = .75) +
  labs(x = "", y = "Proportion of queries") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.01))) +
  scale_fill_manual(values = c("unclassified" = "grey65", color_kingdoms)) +
  thm +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
```

## D - Taxonomic resolution

Resolution is what makes the comparison with the reference annotation
interpretable: the raw count of labelled transcripts says nothing about how
deeply they were labelled.

```{r resolution}
# the four databases use different rank vocabularies (nr carries the full NCBI
# set, RepDB a compact 7-rank taxdump, the reference a PhyloDB-derived scheme).
# Fold
# every label onto the seven canonical ranks before comparing.
CANON <- c(
  superkingdom = "domain", domain = "domain", kingdom = "kingdom",
  phylum = "phylum", subphylum = "phylum",
  superclass = "class", class = "class", subclass = "class", infraclass = "class",
  superorder = "order", order = "order", suborder = "order", infraorder = "order",
  superfamily = "family", family = "family", subfamily = "family",
  genus = "genus", subgenus = "genus",
  `species group` = "species", species = "species", subspecies = "species",
  strain = "species", forma = "species"
)
RANK_LEVELS <- c("species", "genus", "family", "order", "class", "phylum",
                 "kingdom", "domain", "unresolved")

canonise <- function(x) ifelse(x %in% names(CANON), unname(CANON[x]), "unresolved")

# n_direct in the report = transcripts assigned exactly at that taxon, so the
# aggregate report reproduces the per-transcript rank profile without needing
# the (sometimes stale) _lca.tsv
db_profile <- reports %>%
  filter(db %in% REPORT_DBS) %>%
  mutate(cr = canonise(rank)) %>%
  group_by(db, cr) %>% summarise(n = sum(n_direct), .groups = "drop")

# the reference uses a PhyloDB-derived scheme: supergroup and division have no exact
# NCBI equivalent, so they are folded onto kingdom and phylum. family / genus /
# species mean the same thing in both, which is where the comparison is safe.
ref_profile <- ref %>%
  mutate(cr = case_when(
    ref_rank %in% c("species", "genus", "family", "order", "class") ~ ref_rank,
    ref_rank == "division"   ~ "phylum",
    ref_rank == "supergroup" ~ "kingdom",
    TRUE ~ "unresolved"
  )) %>%
  count(cr) %>% mutate(db = REF)

profile <- bind_rows(db_profile, ref_profile) %>%
  mutate(cr = factor(cr, levels = RANK_LEVELS)) %>%
  group_by(db) %>% arrange(cr, .by_group = TRUE) %>%
  mutate(pct = n / sum(n) * 100, cum = cumsum(pct)) %>%
  ungroup()

res_tbl <- profile %>%
  filter(cr != "unresolved") %>%
  select(db, cr, cum) %>%
  pivot_wider(names_from = db, values_from = cum) %>%
  rename(`Rank or finer` = cr)

show_tbl(res_tbl, "Cumulative percentage of the query set assigned at each rank *or finer*. Read down: the reference labels almost everything, but very little of it deeply.")

pv <- function(d, r) {
  v <- profile$cum[profile$db == d & profile$cr == r]
  if (length(v) == 0) NA else v
}
N$fam_repdb <- pv("repdb", "family")
N$fam_nr    <- pv("nr", "family")
N$fam_ref   <- pv(REF, "family")
N$sp_repdb  <- pv("repdb", "species")
N$sp_nr     <- pv("nr", "species")
N$sp_ref    <- pv(REF, "species")
N$gen_repdb <- pv("repdb", "genus")
N$gen_ref   <- pv(REF, "genus")
N$any_repdb <- pv("repdb", "domain")
N$any_nr    <- pv("nr", "domain")
N$any_ref   <- pv(REF, "kingdom")

# Plot only species..phylum. Above phylum the schemes stop corresponding -
# the reference's "supergroup" is not an NCBI kingdom and the mmseqs databases emit
# superkingdom but never kingdom - so a line through those ranks would compare
# vocabulary, not performance. The "any label at all" figures are in the table.
PLOT_RANKS <- rev(c("species", "genus", "family", "order", "class", "phylum"))

plot_res <- profile %>%
  filter(cr %in% PLOT_RANKS) %>%
  mutate(
    cr = factor(cr, levels = PLOT_RANKS),
    db = factor(db, levels = DB_LEVELS)
  ) %>%
  ggplot(aes(as.integer(cr), cum, color = db, group = db)) +
  geom_line(linewidth = .6, key_glyph = "point") +
  geom_point(size = 1.5) +
  scale_x_continuous(
    breaks = seq_along(PLOT_RANKS), labels = str_to_title(PLOT_RANKS),
    expand = expansion(mult = c(0.04, 0.04))
  ) +
  scale_y_continuous(limits = c(0, 100), expand = expansion(mult = c(0, 0.02))) +
  scale_db() +
  labs(x = "Assigned at rank or finer", y = "% of queries") +
  thm +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
```

At **family level or finer** — the deepest rank where the reference's and the
NCBI/UniEuk vocabularies mean the same thing — RepDB places **`r fmt_p(N$fam_repdb)`** of the
queries, nr `r fmt_p(N$fam_nr)`, and the reference annotation only
`r fmt_p(N$fam_ref)`.

## Assembled figure

```{r fig5, fig.width=7.5, fig.height=6.19}
# same shape as the original figure - the two cost/quality panels stacked on the
# left, the taxonomy panels to the right - with A and B given less width now that
# there are four panels. A and B carry no key of their own; the collected legends
# (kingdoms from C, databases from D) sit at the right of the whole assembly.
#
# Built as ONE flat 2x2 grid, not as `(A / B) | (C / D)`: nesting makes two
# independent column groups, patchwork sizes their rows separately, and the D
# tag ends up lower than B. wrap_plots fills row-major, so the plots go in
# reading order and tag_levels spells out the letters they should carry.
#
# free(..., "space", "b") lets panel A keep its own bottom furniture. In a flat
# grid the axis-label row is shared across a row, so C's angled tick labels
# would otherwise size it and strand A's "Peak memory (GB)" title an inch below
# its own ticks.
#
# Geometry: 7.5in wide, which is what panel B needs for the outermost "100" of
# both facets to clear the panel border. The height is solved, not guessed - the
# panels are measured off the rendered PNG and 6.19in is what reproduces the
# panel height of the earlier 7.2x5.94 draft, the flat grid having given ~6% of
# it back to the now-shared axis rows. The original 7.5x5 was close to 3:2, which
# squeezed each panel to ~2.4in tall and forced 8pt type; the panels are now
# nearly square, which is also the right shape for D (a rank profile read left to
# right against a full 0-100 axis).
plot_comp <- wrap_plots(
    free(plot_bench, type = "space", side = "b"), plot_outcome,
    plot_dmnd, plot_res,
    ncol = 2
  ) +
  plot_layout(widths = c(1.3, 1), guides = "collect") +
  # Both collected keys sit in a column on the right. legend.spacing.y is the
  # only lever ggplot gives for spreading stacked keys - there is no
  # space-between - and 40mm is tuned so the outcome key lands beside panel C
  # and the database key beside panel D rather than both bunching at the
  # vertical centre. It is tied to the 6.19in height below; re-tune it if that
  # changes.
  #
  # This has to go through plot_annotation(theme = ), NOT `& theme(...)`: `&`
  # applies to every panel and would overwrite the legend.position = "none" that
  # A and B set, so the figure would come back with three redundant database
  # keys instead of one.
  plot_annotation(
    # wrap_plots filled row-major (bench, outcome, dmnd, res), so the tags are
    # given explicitly rather than left to run A-B-C-D down that order
    tag_levels = list(c("A", "C", "B", "D")),
    theme = theme(legend.position = "right", legend.box = "vertical",
                  legend.box.just = "left", legend.margin = margin(0, 0, 0, 0),
                  legend.box.spacing = unit(3, "mm"),
                  legend.spacing.y = unit(45, "mm"))
  )

saveRDS(plot_comp, fig("comparison.Rds"))
ggsave(fig("Fig5.pdf"), plot_comp, width = 8, height = 6)
plot_comp
```

*A) Wall time and peak memory of the search on `r fmt_n(N$n_queries)` predicted
proteins; the grey line joins each database to its clustered version. B) Identity
and query coverage of the best DIAMOND hit per query. C) Outcome of every query
`r REF` placed in a domain: whether the database agreed with the reference
at domain level, assigned a different domain, found a homolog without resolving
an LCA, or found no homolog at all. D) Cumulative percentage of queries assigned at each rank or finer,
over the range where the reference, NCBI and UniEuk/GTDB rank vocabularies
correspond.*

# Why the reference leaves fewer transcripts unclassified

The reference annotation labels more transcripts than RepDB does, which invites
the reading that RepDB's downsampling has removed sequences needed to annotate
environmental data. It has not: RepDB is ~3.5x smaller than nr and still finds
a homolog for more of the queries (`r fmt_p(N$hit_repdb)` against
`r fmt_p(N$hit_nr)`, table in section B above). The difference is in what
"classified" means on each side, and the sections below take that apart.

## 1. What "unclassified" actually means

`unclassified` conflates two very different failures: *no homolog exists in the
database* (a genuine coverage gap) and *homologs exist but the LCA declined to
commit to a taxon* (a property of the LCA over a taxonomically broad database).
Splitting them separates a genuine coverage gap from an algorithmic artifact.

```{r n-unclassified-reason}
UNINFORMATIVE <- c("unclassified", "root", "cellular organisms")

has_hit <- hits %>% distinct(db, qseqid) %>%
  mutate(db = as.character(db), has_hit = TRUE)

reason <- lca %>%
  mutate(unresolved = is.na(classification) | classification %in% UNINFORMATIVE) %>%
  left_join(has_hit, by = c("db", "seq" = "qseqid")) %>%
  mutate(has_hit = replace_na(has_hit, FALSE))

reason_tbl <- reason %>%
  filter(unresolved) %>%
  count(db, has_hit) %>%
  group_by(db) %>%
  mutate(`% of unresolved` = n / sum(n) * 100) %>%
  ungroup() %>%
  transmute(db,
    Reason = if_else(has_hit, "Homolog found, LCA did not commit", "No homolog in the database"),
    Transcripts = n, `% of unresolved`
  )

N$lca_artifact_pct <- reason_tbl$`% of unresolved`[
  reason_tbl$db == "repdb" & str_starts(reason_tbl$Reason, "Homolog")]

show_tbl(reason_tbl, "Why a transcript ends up without a taxonomic label. DIAMOND and MMseqs2 use different thresholds, so 'homolog found' is a proxy.")
```

For RepDB, **`r fmt_p(N$lca_artifact_pct)` of unresolved transcripts do have a
homolog** — the LCA simply spans too many taxa to commit. That is a property of
the LCA over a tree-of-life database, not a missing-sequence problem.

## 2. Resolution, not coverage

The reference labels more transcripts, but far more shallowly. It comes from a
protist-focused reference set and a different LCA procedure, and its bar for
calling something "classified" is much lower than ours.

```{r n-resolution-headline}
show_tbl(
  res_tbl %>% filter(`Rank or finer` %in% c("species", "genus", "family", "class")),
  "Cumulative % of queries assigned at rank or finer."
)
```

```{r n-resolution-text, echo=FALSE, results="asis"}
cat(sprintf(
  "The reference assigns *some* label to %s of transcripts against RepDB's %s. But it reaches species for only %s (RepDB %s), genus for %s (RepDB %s) and family for %s (RepDB %s). Below family the ordering reverses and stays reversed: the reference's advantage is entirely in shallow labels.\n",
  fmt_p(N$any_ref), fmt_p(N$any_repdb),
  fmt_p(N$sp_ref), fmt_p(N$sp_repdb),
  fmt_p(N$gen_ref), fmt_p(N$gen_repdb),
  fmt_p(N$fam_ref), fmt_p(N$fam_repdb)
))
```

Two further caveats on the reference labels themselves:

```{r n-ref-quality}
ref_counts <- c(N$n_ref_total, N$n_ref_ambiguous, N$n_ref_unclassified)

show_tbl(
  tibble(
    Property = c("Transcripts annotated", "Flagged `ambiguous` by the original study",
                 "Left unclassified by the original study"),
    Transcripts = ref_counts,
    `%` = ref_counts / N$n_ref_total * 100
  ),
  "Quality of the reference annotation used as ground truth."
)
```

## 3. Where RepDB and the reference disagree

```{r n-agreement, eval=("repdb" %in% LCA_DBS)}
agr_tbl <- agreement %>%
  group_by(db) %>%
  summarise(
    `Reference-placed queries` = n(),
    `Also placed by the DB` = sum(!is.na(db_domain)),
    Agree = sum(!is.na(db_domain) & db_domain == ref_domain),
    `% agreement where both assign` = Agree / `Also placed by the DB` * 100,
    .groups = "drop"
  )

N$domain_agreement <- agr_tbl$`% agreement where both assign`[agr_tbl$db == "repdb"]

show_tbl(agr_tbl, "Domain-level agreement with Cohen2024 (Fig. 5C). Domain is the deepest rank at which the reference, NCBI and GTDB nomenclature are directly comparable - below it, RepDB's GTDB prokaryotic names are simply different strings from the reference's NCBI names, so a string comparison would measure nomenclature, not accuracy.")
```

Where both commit to a domain they agree
`r if ("repdb" %in% LCA_DBS) fmt_p(N$domain_agreement) else "-"` of the time, so
the extra transcripts RepDB resolves are not being bought with wrong calls.

## Is the gap concentrated in the clades RepDB caps?

This is the one place where the downsampling does show a cost. The caps in
`config/repdb.yaml` are Opisthokonta
(class, n=20), Ciliophora (order, n=20) and Embryophyta (family, n=20).

```{r n-capped, eval=("repdb" %in% LCA_DBS)}
CAPPED <- c("Opisthokonta", "Ciliophora", "Embryophyta")

gap <- reason %>%
  filter(db == "repdb") %>%
  inner_join(ref, by = "seq") %>%
  filter(!is.na(ref_taxon), ref_taxon != "unclassified")

N$gap_total <- sum(gap$unresolved)
N$gap_pct <- mean(gap$unresolved) * 100

capped_tbl <- map_dfr(CAPPED, function(cl) {
  g <- gap %>% mutate(inside = str_detect(replace_na(ref_lineage, ""), cl))
  tibble(
    Clade = cl,
    `Queries in clade` = sum(g$inside),
    `% unresolved inside` = mean(g$unresolved[g$inside]) * 100,
    `% unresolved outside` = mean(g$unresolved[!g$inside]) * 100
  )
}) %>% mutate(Enrichment = `% unresolved inside` / `% unresolved outside`)

N$opis_in  <- capped_tbl$`% unresolved inside`[capped_tbl$Clade == "Opisthokonta"]
N$opis_out <- capped_tbl$`% unresolved outside`[capped_tbl$Clade == "Opisthokonta"]
N$capped_cost <- sum(capped_tbl$`Queries in clade` * capped_tbl$`% unresolved inside` / 100)

show_tbl(capped_tbl, "Are the transcripts RepDB fails to resolve concentrated in the downsampled clades?", digits = 2)
```


# Session info

```{r session-info}
sessionInfo()
```
