---
title: "RepDB EDA"
date: "`r format(Sys.time(), '%d-%b-%Y')`"
author: "Giacomo Mutti"
output:
    rmdformats::html_clean:
        number_sections: false
        code_folding: hide
        self_contained: true
        fig_caption: true
        gallery: true
        toc_depth: 3
        highlight: kate
---

<!--
STEP 2 of 2. Runs on a laptop. Reads ONLY the .RData written by eda_prepare.Rmd
(step 1, on the HPC) and produces every table and figure. It never touches
results/, /gpfs or the trees.

  Rscript -e 'root <- getwd()
  dir.create(file.path(root, "results/qc"), recursive = TRUE, showWarnings = FALSE)
  rmarkdown::render("workflow/notebooks/eda_report.Rmd",
                    knit_root_dir = root,
                    output_dir    = file.path(root, "results/qc"),
                    output_file   = "eda.html")'

Bind the root to a variable first: render() does setwd(dirname(input)) before it
forces `output_file`, so a lazily evaluated getwd() there resolves against
workflow/notebooks/ instead of the repo root.

If the bundle lives elsewhere, edit `P$bundle` in the `config` chunk below.

Sections whose inputs were missing on the HPC, or whose plotting packages are
missing here, are skipped with a note instead of aborting.
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, cache = FALSE, warning = FALSE, message = FALSE
)
if (nzchar(Sys.which("pdfcrop"))) {
  knitr::knit_hooks$set(crop = knitr::hook_pdfcrop)
  knitr::opts_chunk$set(crop = TRUE)
}
```

```{r libs}
suppressPackageStartupMessages({
  library(tidyverse)
  library(patchwork)
})
theme_set(theme_classic())

.src <- function(p) source(if (file.exists(p)) p else file.path("..", "..", p))
.src("workflow/scripts/functions.R")
.src("workflow/scripts/palettes.R")

# the tree and treemap panels need packages that may not be installed on a
# laptop; everything else works without them
pkg <- function(...) all(vapply(c(...), requireNamespace, logical(1), quietly = TRUE))
HAVE_TREE_PKGS <- pkg("ggtree", "ggtreeExtra", "ggnewscale")
HAVE_TREEMAP   <- pkg("WeightedTreemaps")
if (HAVE_TREE_PKGS) suppressPackageStartupMessages({
  library(ggtree); library(ggtreeExtra)
})
if (HAVE_TREEMAP) suppressPackageStartupMessages(library(WeightedTreemaps))
```

```{r config}
P <- list(
  bundle         = "results/qc/eda_bundle.RData",
  figdir         = "results/plots",
  comparison_rds = "results/plots/comparison.Rds",
  recall_rds     = "results/plots/recall.RDS"
)
```

```{r load-bundle}
.rp <- function(p) if (!file.exists(p) && file.exists(file.path("..", "..", p))) file.path("..", "..", p) else p
bundle_path <- .rp(P$bundle)
if (!file.exists(bundle_path)) {
  stop("bundle not found: ", bundle_path,
       "\nRun workflow/notebooks/eda_prepare.Rmd on the HPC and copy the .RData here,",
       "\nor edit P$bundle in the config chunk above.")
}
load(bundle_path)

figdir <- .rp(P$figdir)
dir.create(figdir, showWarnings = FALSE, recursive = TRUE)
fig <- function(name) file.path(figdir, name)

# taxonomy lookup, rebuilt from the per-organism table
tax <- repdb_df %>% select(mnemo, k, p, c, o, f, g, s)

TREES <- HAVE_TREES && HAVE_TREE_PKGS
skip_note <- function(cond, msg) if (!cond) cat("> **Skipped**: ", msg, "\n\n", sep = "")
```

```{r provenance, echo=FALSE}
show_meta <- tibble(
  Field = c("Bundle", "Created", "Host", "R", "Git commit", "Organisms",
            "Trees", "nr tables", "E. coli ids", "contaminants.txt"),
  Value = c(
    bundle_path,
    format(bundle_meta$created, "%Y-%m-%d %H:%M"),
    bundle_meta$host, bundle_meta$r_version,
    paste(bundle_meta$git_commit, collapse = ""),
    format(bundle_meta$n_organisms, big.mark = ","),
    as.character(HAVE_TREES), as.character(HAVE_NR),
    as.character(HAVE_ECOLI), as.character(HAVE_CONTAM_FILE)
  )
)
knitr::kable(show_meta, caption = "Where this data came from.")
```

```{r reporting-helpers}
# N collects every quantity quoted in the manuscript so the inline text, the
# tables and the final checklist can never drift apart.
N <- list()

show_tbl <- function(df, caption = NULL, digits = 2) {
  k <- knitr::kable(df, caption = caption, digits = digits,
                    format.args = list(big.mark = ","))
  if (requireNamespace("kableExtra", quietly = TRUE)) {
    k <- kableExtra::kable_styling(k,
      bootstrap_options = c("striped", "condensed"), full_width = FALSE)
  }
  k
}

fmt_n <- function(x) formatC(round(x), format = "d", big.mark = ",")
fmt_m <- function(x, d = 1) paste0(formatC(x / 1e6, format = "f", digits = d), "M")
fmt_b <- function(x, d = 2) paste0(formatC(x / 1e9, format = "f", digits = d), "B")
fmt_p <- function(x, d = 1) paste0(formatC(x, format = "f", digits = d), "%")
fmt_d <- function(x, d = 1) formatC(x, format = "f", digits = d)
fmt_gb <- function(x, d = 0) paste0(formatC(x / 2^30, format = "f", digits = d), "GB")
fmt_msd <- function(m, s, d = 1) sprintf("%s±%s", fmt_d(m, d), fmt_d(s, d))
```

# General statistics of RepDB

## Composition

```{r n-composition}
composition <- repdb_df %>%
  group_by(Kingdom = k) %>%
  summarise(
    Proteomes = n_distinct(mnemo),
    Species   = n_distinct(s, na.rm = TRUE),
    `Phyla/Classes` = n_distinct(label, na.rm = TRUE),
    Sequences = sum(num_seqs, na.rm = TRUE),
    `Sequences (M)` = sum(num_seqs, na.rm = TRUE) / 1e6,
    `Completeness mean` = mean(completeness, na.rm = TRUE),
    `Completeness sd`   = sd(completeness, na.rm = TRUE),
    `Contamination mean` = mean(contamination, na.rm = TRUE),
    `Contamination sd`   = sd(contamination, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(Sequences))

N$n_prot  <- n_distinct(repdb_df$mnemo)
N$n_seqs  <- sum(repdb_df$num_seqs, na.rm = TRUE)
N$n_sp    <- n_distinct(repdb_df$s, na.rm = TRUE)
N$n_clust <- sum(clust_stats$num_seqs, na.rm = TRUE)

kv <- function(kingdom, col) composition[[col]][composition$Kingdom == kingdom]
for (kk in c("Bacteria", "Archaea", "Eukaryota", "Viruses")) {
  N[[paste0("n_prot_", kk)]] <- kv(kk, "Proteomes")
  N[[paste0("n_seqs_", kk)]] <- kv(kk, "Sequences")
  N[[paste0("compl_", kk)]]  <- fmt_msd(kv(kk, "Completeness mean"), kv(kk, "Completeness sd"))
  N[[paste0("cont_", kk)]]   <- fmt_msd(kv(kk, "Contamination mean"), kv(kk, "Contamination sd"))
}

composition %>%
  bind_rows(summarise(composition,
    Kingdom = "TOTAL",
    across(c(Proteomes, Species, `Phyla/Classes`, Sequences, `Sequences (M)`), sum)
  )) %>%
  show_tbl("RepDB composition per kingdom. Completeness is BUSCO (eukaryotes, lineage-specific ortholog sets) or CheckM2 (prokaryotes); contamination is the mmseqs cross-kingdom flagging (eukaryotes) or CheckM2 (prokaryotes).")
```

The final database includes **`r fmt_m(N$n_seqs, 0)`** sequences
(`r fmt_n(N$n_seqs)`) from `r fmt_n(N$n_prot)` proteomes, with a marked
predominance of bacterial proteins
(**`r fmt_m(N$n_seqs_Bacteria, 0)` across `r fmt_n(N$n_prot_Bacteria)` proteomes**),
followed by Eukaryotes (`r fmt_m(N$n_seqs_Eukaryota, 1)` in
`r fmt_n(N$n_prot_Eukaryota)` proteomes), Archaea
(`r fmt_m(N$n_seqs_Archaea, 0)`, `r fmt_n(N$n_prot_Archaea)` proteomes) and
Viruses (`r fmt_m(N$n_seqs_Viruses, 1)`, `r fmt_n(N$n_prot_Viruses)` proteomes).

Completeness and contamination are homogeneous in Bacteria
(`r N$compl_Bacteria`%, `r N$cont_Bacteria`%; mean ± SD) and Archaea
(`r N$compl_Archaea`%, `r N$cont_Archaea`%), and vary substantially across
eukaryotic phyla (`r N$compl_Eukaryota`%, `r N$cont_Eukaryota`%).

## Most abundant clades

```{r n-clades}
clade_tbl <- repdb_df %>%
  filter(!is.na(label)) %>%
  group_by(Kingdom = k, `Phylum/Class` = label) %>%
  summarise(
    Proteomes = n_distinct(mnemo),
    Species   = n_distinct(s, na.rm = TRUE),
    Sequences = sum(num_seqs, na.rm = TRUE),
    `Sequences (M)` = sum(num_seqs, na.rm = TRUE) / 1e6,
    .groups = "drop"
  ) %>%
  mutate(`% of RepDB` = Sequences / N$n_seqs * 100) %>%
  arrange(desc(Sequences))

N$top_clades <- paste(head(clade_tbl$`Phylum/Class`, 3), collapse = ", ")
N$top_clade_prop <- clade_tbl$`% of RepDB`[1]

show_tbl(head(clade_tbl, 25),
  "Sequences and proteomes per clade - the values labelled on the backbone tree (Fig. 2A). Top 25; the full table drives the figure.")
```

The most abundant clades are **`r N$top_clades`**; the first alone accounts for
`r fmt_p(N$top_clade_prop)` of RepDB.

## Data types and assembly levels

```{r n-prokaryote-types}
proka_type <- repdb_df %>%
  filter(k %in% c("Bacteria", "Archaea")) %>%
  count(`Assembly level` = type) %>%
  mutate(`%` = n / sum(n) * 100) %>%
  arrange(desc(`%`))

N$proka_top_type <- proka_type$`Assembly level`[1]
N$proka_top_prop <- proka_type$`%`[1]

show_tbl(proka_type, "Assembly level of the prokaryotic (GTDB) genomes.")
```

All prokaryotic entries come from genomic assemblies, mostly at the
**`r str_to_lower(N$proka_top_type)`** level (`r fmt_p(N$proka_top_prop)` of the
cases).

```{r n-eukaryote-types}
euk_type_tbl <- repdb_df %>%
  filter(k == "Eukaryota") %>%
  group_by(`Data type` = type) %>%
  summarise(Proteomes = n(), Sequences = sum(num_seqs, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    `% proteomes` = Proteomes / sum(Proteomes) * 100,
    `% sequences` = Sequences / sum(Sequences) * 100
  ) %>%
  arrange(desc(Sequences))

# "transcriptomic data" = Transcriptome + S.C. transcriptome (EST reported apart)
transcr <- repdb_df %>%
  filter(k == "Eukaryota") %>%
  mutate(Origin = case_when(
    grepl("transcriptome", type, ignore.case = TRUE) ~ "transcriptome-derived",
    type == "EST" ~ "EST",
    TRUE ~ "genome-derived"
  )) %>%
  group_by(Origin) %>%
  summarise(Proteomes = n(), Sequences = sum(num_seqs, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    `% proteomes` = Proteomes / sum(Proteomes) * 100,
    `% sequences` = Sequences / sum(Sequences) * 100
  )

pick <- function(df, origin, col) {
  v <- df[[col]][df$Origin == origin]
  if (length(v) == 0) 0 else v
}
N$transcr_seq_prop  <- pick(transcr, "transcriptome-derived", "% sequences")
N$transcr_prot_prop <- pick(transcr, "transcriptome-derived", "% proteomes")

show_tbl(euk_type_tbl, "Eukaryotic proteomes and sequences per data type.")
show_tbl(transcr, "Transcriptome- vs. genome-derived eukaryotic proteomes.")
```

Eukaryotic proteins are mostly derived from transcriptomic data
(**`r fmt_p(N$transcr_seq_prop)` of the proteins**,
`r fmt_p(N$transcr_prot_prop)` of the proteomes).

## Database footprint

```{r n-footprint}
if (!is.null(footprint) && nrow(footprint)) {
  N$size_fasta <- footprint$Bytes[footprint$Artifact == "FASTA (gzipped)"]
  show_tbl(select(footprint, Artifact, GB), "Disk footprint of the RepDB artifacts.")
} else {
  cat("> No RepDB artifacts were found when the bundle was built.\n")
}
```

```{r n-buildtime}
if (!is.null(bench_build)) {
  show_tbl(bench_build,
    "Build time and memory per step (snakemake benchmarks): `s` seconds, `h:m:s` wall time, `max_rss` peak RSS in MB.")
} else {
  cat("> No build benchmarks were found when the bundle was built.\n")
}
```

# Taxonomic tree (Fig. 2)

```{r tree-note, echo=FALSE, results="asis"}
skip_note(HAVE_TREES,
  "the reduced backbone trees were missing when the bundle was built. Build them (see the `obtain-reduced-prokatree` chunk in eda_prepare.Rmd) and re-run step 1. All the numbers above and below are unaffected.")
skip_note(!HAVE_TREES || HAVE_TREE_PKGS,
  "the trees are in the bundle but `ggtree` / `ggtreeExtra` / `ggnewscale` are not installed here. `install.packages(\"ggnewscale\")` and `BiocManager::install(c(\"ggtree\", \"ggtreeExtra\"))`.")
```

```{r fig1-params}
tl <- 35.1 #  distance of type barplot
x_lab <- 13.7 # distance of geomlabel
min_x <- -15
offsets <- c(0.75, 0.005) # c(0.575, 0.005)
size_bar <- 2.75
offset_tm <- 0.375
```

```{r fig1-data}
repdb_summary <- repdb_df %>%
  group_by(label) %>%
  summarise(
    n_seqs = sum(num_seqs, na.rm = TRUE),
    n_gen  = n_distinct(mnemo),
    .groups = "drop"
  ) %>%
  mutate(n_seqs_t = signif(n_seqs / 1e6, 2))

type_df <- repdb_df %>%
  count(label, type) %>%
  mutate(type = factor(type, levels = rev(names(c(ass_palette, type_palette))))) %>%
  group_by(label) %>%
  arrange(type, .by_group = TRUE) %>%
  mutate(
    freq = n / sum(n),
    ymax = tl + cumsum(freq),
    ymin = lag(ymax, default = tl)
  ) %>%
  ungroup()
```

```{r fig1, eval=TREES}
p1 <- eukaproka_tree %>%
  fortify() %>%
  left_join(repdb_summary, by = "label") %>%
  ggtree(layout = "circular") +
  geom_tiplab(size = 2.2, hjust = 0, color = "black") +
  geom_label(
    aes(angle = angle, label = n_seqs_t, fill = n_seqs,
        x = max(fortify(eukaproka_tree)$x) + x_lab),
    show.legend = F, size = 1.5, color = "white", hjust = 1, vjust = .5
  ) +
  scale_x_continuous(limits = c(min_x, NA)) +
  scale_fill_viridis_c(trans = "log")

p2 <- p1 +
  ggnewscale::new_scale_fill() +
  geom_fruit(
    data = repdb_df, geom = geom_boxplot,
    mapping = aes(y = label, x = contamination, group = label, fill = k),
    orientation = "y", offset = offsets[1], size = .1,
    outlier.size = 0.2, outlier.stroke = 0.08, pwidth = 0.1,
    axis.params = list(axis = "x", text.size = 1, hjust = 1, vjust = 1, nbreak = 3),
    grid.params = list()
  ) +
  geom_fruit(
    data = repdb_df, geom = geom_boxplot,
    mapping = aes(y = label, x = completeness, group = label, fill = k),
    orientation = "y", offset = offsets[2], size = .1, pwidth = 0.1,
    outlier.size = 0.2, outlier.stroke = 0.08,
    axis.params = list(axis = "x", text.size = 1, hjust = 1, vjust = 1, nbreak = 3),
    grid.params = list()
  ) +
  scale_fill_manual(values = color_kingdoms) +
  ggnewscale::new_scale_fill() +
  geom_fruit(
    mapping = aes(y = label, xmin = ymin, xmax = ymax, color = type),
    size = size_bar, data = type_df, geom = geom_linerange, show.legend = FALSE
  ) +
  scale_color_manual(values = c(ass_palette, type_palette)) +
  theme(legend.position = "none")
```

```{r add-legend, eval=TREES}
p_k <- color_kingdoms %>%
  enframe() %>% mutate(y = 1) %>%
  ggplot(aes(name, y, fill = value)) +
  geom_tile() + scale_fill_identity() + coord_cartesian(ratio = 1) +
  labs(subtitle = "Kingdom") + theme_legend

p_type <- c(type_palette, ass_palette) %>%
  enframe() %>% filter(name != "transcriptome,EST") %>%
  mutate(y = 1, order = 1:nrow(.)) %>%
  ggplot(aes(fct_reorder(name, order), y, fill = value)) +
  geom_tile() + scale_fill_identity() + coord_cartesian(ratio = 1) +
  labs(subtitle = "Data Type") + theme_legend

p3 <- p2 + inset_element(p_k | p_type,
  left = 0.1, bottom = -.05, right = 0.45, top = 0.16,
  align_to = "full", clip = FALSE)
```

```{r add-treemap, eval=TREES}
repdb_counts <- repdb_df %>%
  group_by(k, label) %>%
  summarise(n = sum(num_seqs, na.rm = TRUE), .groups = "drop")

if (HAVE_TREEMAP) {
  tm <- voronoiTreemap(
    data = filter(repdb_counts, n > 10000),
    levels = c("k", "label"), cell_size = "n", shape = "circle", seed = 123
  )
  pdf(NULL)
  drawTreemap(tm,
    label_size = 1, color_palette = color_kingdoms[unique(repdb_counts$k)],
    label_color = "grey90", border_level = c(1, 2),
    border_color = c("black", "grey20"), border_size = c(2, 0.5),
    width = 0.8, height = 0.8)
  p_tm <- grid::grid.grab()
  dev.off()

  p4 <- p3 + inset_element(p_tm,
    left = offset_tm, bottom = offset_tm,
    right = 1 - offset_tm, top = 1 - offset_tm,
    align_to = "full", clip = FALSE) +
    theme(plot.background = element_blank(), panel.background = element_blank())
} else {
  # the Voronoi inset needs WeightedTreemaps; the tree itself does not
  message("WeightedTreemaps not installed - Fig. 2 is written without panel B")
  p4 <- p3
}

```

```{r fig1-show, eval=TREES, fig.width=11, fig.height=11}
p4
```

# Viruses

```{r n-viral-classes}
viral_class <- virus_df %>%
  group_by(`Viral class` = c) %>%
  summarise(
    Genomes   = n(),
    Species   = n_distinct(s, na.rm = TRUE),
    Sequences = sum(num_seqs, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    `% genomes`   = Genomes / sum(Genomes) * 100,
    `% sequences` = Sequences / sum(Sequences) * 100
  ) %>%
  arrange(desc(Sequences))

vc <- function(cls, col) {
  v <- viral_class[[col]][viral_class$`Viral class` == cls]
  if (length(v) == 0) NA else v
}
N$caudo_seq_prop <- vc("Caudoviricetes", "% sequences")
N$caudo_species  <- vc("Caudoviricetes", "Species")
N$mega_seq_prop  <- vc("Megaviricetes", "% sequences")
N$mega_species   <- vc("Megaviricetes", "Species")
N$vir_2nd        <- viral_class$`Viral class`[2]

show_tbl(viral_class,
  "Viral classes in RepDB, ranked by number of sequences (backs the Caudoviricetes / Megaviricetes figures in the text and Fig. S2).")
```

The dsDNA class **Caudoviricetes** is the most represented group
(**`r fmt_p(N$caudo_seq_prop)` of the viral proteins** from
**`r fmt_n(N$caudo_species)` different species**). The next most abundant group,
**`r N$vir_2nd`**, represents `r fmt_p(N$mega_seq_prop)` of the viral proteins
in RepDB despite deriving from just `r fmt_n(N$mega_species)` distinct species,
a result of their large genome sizes.

```{r n-virus-molecule}
show_tbl(
  virus_df %>%
    mutate(Molecule_type = replace_na(Molecule_type, "unknown")) %>%
    group_by(`Molecule type` = Molecule_type) %>%
    summarise(Genomes = n(), Sequences = sum(num_seqs, na.rm = TRUE), .groups = "drop") %>%
    mutate(
      `% genomes` = Genomes / sum(Genomes) * 100,
      `% sequences` = Sequences / sum(Sequences) * 100
    ) %>%
    arrange(desc(Genomes)),
  "Viral genomes per molecule type (the Fig. S2 barplot)."
)
```

## Viral tree (Fig. S2)

```{r virtree, eval=TREES, fig.width=4.3, fig.height=4.5}
virtype_df <- virus_df %>%
  mutate(
    Molecule_type = ifelse(is.na(Molecule_type), "unknown", Molecule_type),
    Molecule_type = factor(Molecule_type, levels = names(virus_colors))
  ) %>%
  group_by(label, Molecule_type) %>%
  summarise(n = n(), .groups = "drop_last") %>%
  mutate(freq = n / sum(n)) %>%
  ungroup()

pvir1 <- virtree %>%
  fortify() %>%
  left_join(repdb_summary, by = "label") %>%
  ggtree(size = .2) +
  geom_tiplab(size = 2, hjust = 0, color = "grey20") +
  geom_nodelab(size = 1, color = "firebrick", hjust = 1.1, vjust = -.7) +
  geom_label(
    aes(label = n_seqs, fill = n_seqs, x = max(fortify(virtree)$x + 3)),
    show.legend = F, size = 1.5, color = "white", hjust = 1, vjust = .5
  ) +
  scale_fill_viridis_c(trans = "log")

pvir2 <- pvir1 +
  ggnewscale::new_scale_fill() +
  geom_fruit(
    data = virtype_df, geom = geom_bar, stat = "identity",
    mapping = aes(y = label, x = freq, group = label, fill = Molecule_type),
    orientation = "y", offset = .77, pwidth = .5,
    axis.params = list(axis = "x", text.size = 1, hjust = 1, vjust = 1, nbreak = 3,
      title = "Virus Type", title.size = 2, title.height = 0.04, title.color = "grey10")
  ) +
  geom_fruit(
    data = filter(repdb_df, k == "Viruses"), geom = geom_boxplot,
    mapping = aes(y = label, x = log10(num_seqs), group = label),
    fill = color_kingdoms["Viruses"], orientation = "y", offset = 0.05,
    size = .1, outlier.size = 0.2, outlier.stroke = 0.08, pwidth = 1,
    axis.params = list(axis = "x", text.size = 1, hjust = 1, vjust = 1, nbreak = 5,
      title = "log10(# Seqs)", title.size = 2, title.height = 0.04, title.color = "grey10"),
    grid.params = list()
  ) +
  scale_fill_manual(values = virus_colors) +
  labs(fill = "Molecule Type") +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 5),
    legend.key.size = unit(2, "mm"),
    legend.text = element_text(size = 4, margin = margin(l = 0.1)),
    legend.key.spacing.y = unit(0, "mm"),
    legend.box.spacing = unit(0, "mm")
  )

pvir2
```

# Decontamination of eukaryotic proteomes (Fig. 3)

```{r euk-df}
euk_df <- repdb_df %>%
  filter(k == "Eukaryota") %>%
  mutate(
    label = factor(label, levels = rev(p_order)),
    db = factor(db, levels = c("custom", "uniprot", "eukprot", "p10k"))
  ) %>%
  replace_na(list(contamination = 0))

N$n_euk_prot <- nrow(euk_df)
N$n_euk_seqs <- sum(euk_df$num_seqs, na.rm = TRUE)
```

We evaluated the decontamination module across the **`r fmt_n(N$n_euk_prot)`**
eukaryotic proteomes selected for RepDB.

```{r n-contam-source}
show_tbl(contam_check, "Source of the contaminant counts. The two rows should agree; if `repdb_meta.tsv` reports zero, it was written before the organism-id fix in `analyze_stats.R` and needs `make_db_meta` re-running. Step 1 uses the primary artifact either way.")
```

```{r n-contam-warn, echo=FALSE, results="asis"}
n_file <- contam_check$`Flagged sequences`[1]
n_meta <- contam_check$`Flagged sequences`[2]
if (!HAVE_CONTAM_FILE) {
  cat(sprintf("> **Note**: `%s` was not found when the bundle was built, so contamination falls back to the `repdb_meta.tsv` tally.\n\n", contam_file))
} else if (isTRUE(n_meta == 0)) {
  cat("> **Note**: `repdb_meta.tsv` reports zero contaminants for every organism - the symptom of the organism-id bug. The figures below are recomputed from `contaminants.txt` and are correct, but re-run `make_db_meta` so the published provenance table matches.\n\n")
} else if (isTRUE(n_file != n_meta)) {
  cat(sprintf("> **Note**: the two contaminant tallies disagree (%s vs %s). Using `contaminants.txt`.\n\n",
              fmt_n(n_file), fmt_n(n_meta)))
}
if (sum(euk_df$contamination, na.rm = TRUE) == 0) {
  cat("> **Warning**: every eukaryotic proteome has 0% contamination. That is not a plausible result - check `contaminants.txt` and the sequence-id format before using anything below.\n\n")
}
```

## Contamination per source database (Fig. 3A)

```{r n-cont-db}
cont_db <- euk_df %>%
  group_by(db) %>%
  summarise(
    n_prot  = n(),
    n_conts = sum(n_conts, na.rm = TRUE),
    n_tot   = sum(num_seqs, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(prop = n_conts / n_tot * 100)

dbv <- function(d, col) {
  v <- cont_db[[col]][as.character(cont_db$db) == d]
  if (length(v) == 0) NA else v
}
for (d in c("eukprot", "p10k", "uniprot", "custom")) {
  N[[paste0("cont_prop_", d)]] <- dbv(d, "prop")
}
N$n_conts_total <- sum(cont_db$n_conts, na.rm = TRUE)

show_tbl(
  cont_db %>%
    transmute(Source = db, Proteomes = n_prot, Sequences = n_tot,
              Contaminants = n_conts, `% contaminated` = prop) %>%
    arrange(desc(`% contaminated`)),
  "Flagged cross-kingdom contaminants per source database (Fig. 3A)."
)
```

EukProt shows the highest fraction of putative contaminant sequences
(**`r fmt_p(N$cont_prop_eukprot, 2)`**), P10K a lower one
(**`r fmt_p(N$cont_prop_p10k, 2)`**) and UniProt the lowest
(**`r fmt_p(N$cont_prop_uniprot, 2)`**). In total `r fmt_n(N$n_conts_total)`
eukaryotic sequences are flagged.

## Contamination per class (Fig. 3B)

```{r n-cont-class}
cont_class <- euk_df %>%
  group_by(Class = as.character(label)) %>%
  summarise(
    Proteomes = n(),
    `Median % contaminated` = median(contamination, na.rm = TRUE),
    `Mean % contaminated`   = mean(contamination, na.rm = TRUE),
    `Max % contaminated`    = max(contamination, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(`Median % contaminated`))

N$worst_class     <- cont_class$Class[1]
N$worst_class_med <- cont_class$`Median % contaminated`[1]

show_tbl(head(cont_class, 25), "Eukaryotic classes ranked by median contamination (top 25).")
```

**`r N$worst_class`** is the most affected taxon, reaching a median
contamination of **`r fmt_p(N$worst_class_med, 2)`**.

## Contamination per data type

```{r n-cont-type}
cont_type <- euk_df %>%
  group_by(`Data type` = type) %>%
  summarise(
    Proteomes = n(),
    `Mean % contaminated`   = mean(contamination, na.rm = TRUE),
    `Median % contaminated` = median(contamination, na.rm = TRUE),
    `Max % contaminated`    = max(contamination, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(`Mean % contaminated`))

tv <- function(ty) {
  v <- cont_type$`Mean % contaminated`[cont_type$`Data type` == ty]
  if (length(v) == 0) NA else v
}
N$cont_transcriptome    <- tv("Transcriptome")
N$cont_genome           <- tv("Genome")
N$cont_sc_transcriptome <- tv("S.C. transcriptome")
N$cont_sc_genome        <- tv("S.C. genome")

show_tbl(cont_type, "Contamination per data type (Fig. 3B colouring).")
```

Transcriptome-derived proteomes show higher contamination than genomic
resources (**`r fmt_p(N$cont_transcriptome, 2)`** versus
**`r fmt_p(N$cont_genome, 2)`** mean contamination); the trend reverses for
single-cell approaches (`r fmt_p(N$cont_sc_transcriptome, 2)` for S.C.
transcriptomes versus `r fmt_p(N$cont_sc_genome, 2)` for S.C. genomes).

## Highly contaminated proteomes

```{r n-cont-thresholds}
thr <- c(1, 2.5, 5, 10, 20, 50)
cont_thr <- tibble(
  Threshold = paste0(">= ", thr, "%"),
  Proteomes = map_int(thr, ~ sum(euk_df$contamination >= .x, na.rm = TRUE))
) %>%
  mutate(`% of eukaryotic proteomes` = Proteomes / N$n_euk_prot * 100)

N$n_cont5  <- cont_thr$Proteomes[thr == 5]
N$n_cont25 <- cont_thr$Proteomes[thr == 2.5]
N$n_cont20 <- cont_thr$Proteomes[thr == 20]

show_tbl(cont_thr, "Eukaryotic proteomes above each contamination threshold.")
```

```{r n-cont20-breakdown}
above20 <- euk_df %>% filter(contamination > 20)

N$n_above20 <- nrow(above20)
N$n_above20_genome <- sum(grepl("genome", above20$type, ignore.case = TRUE))

show_tbl(
  above20 %>%
    arrange(desc(contamination)) %>%
    transmute(ID = mnemo, Source = as.character(db), Species = s,
              Class = as.character(label), `Data type` = type,
              Sequences = num_seqs, Contaminants = n_conts,
              `% contaminated` = contamination),
  "Every eukaryotic proteome with more than 20% flagged sequences."
)
```

```{r n-cont20-text, echo=FALSE, results="asis"}
if (N$n_above20 > 0) {
  w <- which.max(above20$contamination)
  cat(sprintf(
    "**%s** proteomes have more than 20%% contamination, of which **%s** derive from genome assemblies. The most extreme case is **%s** (*%s*), with %s of its predicted proteins flagged as putative contaminants.\n",
    fmt_n(N$n_above20), fmt_n(N$n_above20_genome),
    above20$mnemo[w], above20$s[w], fmt_p(above20$contamination[w], 1)
  ))
} else {
  cat("No eukaryotic proteome exceeds 20% contamination in this run.\n")
}
```

## Co-occurrence of contaminants (Fig. 3D)

`cooc_by_genus` is one row per (eukaryotic proteome, donor genus), where `n` is
the number of **donor genomes** of that genus sharing a contaminant cluster with
the proteome. It measures how *broadly* a genus co-occurs, and scales with how
many genomes GTDB holds for it — fine for ranking clades within a proteome,
which is what this panel does, but it is not a protein count. The case studies
below use a per-protein attribution for the figures quoted in the text.

```{r co-occurrence}
# proteomes with at least 5% contaminated proteins
interesting <- filter(euk_df, contamination >= 5)$mnemo

top_coocs <- cooc_by_genus %>%
  filter(euka %in% interesting, n > 5) %>%
  group_by(euka) %>%
  mutate(prop = n / sum(n), scaled_prop = scales::rescale(prop)) %>%
  mutate(is_first = dense_rank(desc(n)) == 1) %>%
  ungroup() %>%
  left_join(tax, by = c("euka" = "mnemo"))

top_1_cooc <- top_coocs %>%
  group_by(euka) %>%
  slice_max(order_by = scaled_prop, n = 1, with_ties = F) %>%
  mutate(scaled_prop = 1.01) %>%
  arrange(clade)
```

```{r n-cooc-genera}
N$n_interesting <- length(interesting)

cooc_genera <- top_1_cooc %>%
  ungroup() %>%
  count(`Donor kingdom` = donor_k, Genus = clade, name = "Proteomes") %>%
  arrange(desc(Proteomes)) %>%
  mutate(`% of affected proteomes` = Proteomes / N$n_interesting * 100)

N$top_cooc_genera <- paste0("*", head(cooc_genera$Genus, 3), "*", collapse = ", ")

show_tbl(head(cooc_genera, 20), sprintf(
  "Most common putative contamination source: the genus with the highest co-occurrence in each of the %d eukaryotic proteomes with >=5%% flagged sequences (top 20). Co-occurrence is counted in donor genomes; see the note above.",
  N$n_interesting))
```

At the genus level, **`r N$top_cooc_genera`**, in this order, are the most
common sources of contaminants.

```{r n-cooc-kingdom}
show_tbl(
  top_1_cooc %>%
    ungroup() %>%
    count(`Donor kingdom` = donor_k, name = "Proteomes") %>%
    mutate(`%` = Proteomes / sum(Proteomes) * 100),
  "Kingdom of the top co-occurring contaminant."
)
```

### Case studies

```{r n-case-studies}
# EP01030 = Chromulinospumella sphaerica; P10K-NCBI-001938 = Paulinella micropora
case_tbl <- cooc_case %>%
  group_by(Proteome) %>%
  mutate(`% of flagged` = `Flagged proteins` / sum(`Flagged proteins`) * 100) %>%
  slice_max(`Flagged proteins`, n = 8) %>%
  ungroup() %>%
  arrange(Proteome, desc(`Flagged proteins`))

show_tbl(case_tbl)
```

```{r n-case-text, echo=FALSE, results="asis"}
case_line <- function(id) {
  r <- euk_df %>% filter(mnemo == id)
  top <- case_tbl %>% filter(Proteome == id) %>% slice_head(n = 1)
  if (!nrow(r)) return(invisible(NULL))
  cat(sprintf(
    "- **%s** (*%s*, %s, %s): %s of %s proteins flagged (%s); top putative source *%s* (%s flagged proteins).\n",
    id, r$s[1], r$db[1], r$type[1],
    fmt_n(r$n_conts[1]), fmt_n(r$num_seqs[1]), fmt_p(r$contamination[1], 1),
    if (nrow(top)) top$Genus[1] else NA,
    if (nrow(top)) fmt_n(top$`Flagged proteins`[1]) else NA
  ))
}
walk(unique(cooc_case$Proteome), case_line)
```

```{r cooc-plot}
plot_cooc <- top_coocs %>%
  mutate(s = factor(s, levels = unique(top_1_cooc$s))) %>%
  ggplot(aes(scaled_prop, s)) +
  geom_point(aes(color = is_first), pch = "|") +
  geom_text(aes(label = clade), data = top_1_cooc, hjust = 0, size = 1.5) +
  scale_x_continuous(expand = expansion(add = c(.025, .25)), breaks = seq(0, 1, .25)) +
  scale_color_manual(values = c("grey60", "red")) +
  labs(y = "", x = "Scaled prop. of Co-occurence") +
  theme(legend.position = "none", axis.text.y = element_text(size = 5))
```

## Fig. 3

```{r fig3, fig.width=9, fig.height=6}
plot_cont_db <- cont_db %>%
  ggplot(aes(n_conts, fct_reorder(db, n_conts), fill = db)) +
  geom_col() +
  geom_text(aes(
    label = paste0(round(prop, 2), "%"),
    hjust = ifelse(n_conts > 1e5, 1.1, -0.1),
    color = ifelse(n_conts > 1e5, "white", "black")
  )) +
  labs(x = "# Contaminants", y = "DB") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  scale_color_identity() +
  scale_fill_manual(values = color_dbs) +
  theme(legend.position = "none")

db_order <- cont_db %>% arrange(prop) %>% pull(db) %>% as.character()

plot_type_db <- euk_df %>%
  count(db, type) %>%
  mutate(db = factor(db, levels = db_order)) %>%
  ggplot(aes(n, db, fill = type)) +
  geom_col(position = "fill") +
  labs(x = "% Data Type", y = "") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  scale_fill_manual(values = type_palette) +
  theme(legend.position = "none")

plot_cont_prot <- euk_df %>%
  ggplot(aes(contamination, fct_reorder(label, contamination, .na_rm = T))) +
  geom_boxplot(aes(color = type), size = .2, outlier.size = .1) +
  ggrepel::geom_text_repel(
    data = . %>% filter(contamination > 20),
    aes(label = g), size = 1.5, force = .1, box.padding = 0.1,
    nudge_x = 1, nudge_y = .25,
    segment.curvature = -0.1, segment.ncp = .3, segment.angle = 20
  ) +
  labs(x = "% Contaminated", y = "Eukaryotic class", color = "Data Type") +
  guides(color = guide_legend(ncol = 1, byrow = TRUE,
    override.aes = list(linewidth = .3, fill = "white"))) +
  scale_color_manual(values = type_palette) +
  theme(
    axis.text.y = element_text(size = 7),
    legend.text = element_text(size = 7),
    legend.title = element_text(size = 7),
    legend.key.spacing.y = unit(0, "mm"),
    legend.key.spacing.x = unit(0, "mm"),
    legend.key.size = unit(3, "mm"),
    legend.key = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    legend.position = c(0.75, 0.88)
  )

cont_left  <- (plot_cont_db / plot_cont_prot) + plot_layout(heights = c(1, 5))
cont_right <- (plot_type_db / plot_cooc) + plot_layout(heights = c(1, 5))

plot_cont <- (cont_left | cont_right) + plot_annotation(tag_levels = "A")
plot_cont
```

# Comparison with nr and ClusteredNR (Fig. 4)

```{r nr-note, echo=FALSE, results="asis"}
skip_note(HAVE_NR,
  "the NR taxid count tables were not reachable when the bundle was built. Produce them (see the `count-nr` chunk in eda_prepare.Rmd) and re-run step 1.")
```

```{r nr-derive, eval=HAVE_NR}
# the repdb side only needs the bundle, so it is derived here rather than on the HPC
repdb_classes <- repdb_df %>%
  group_by(k, p, c) %>%
  summarise(n = sum(num_seqs, na.rm = TRUE), .groups = "drop") %>%
  mutate(db = "repdb")

clustrepdb_classes <- clust_stats %>%
  left_join(distinct(tax, k, p, c), by = "c") %>%
  group_by(k, p, c) %>%
  summarise(n = sum(num_seqs), .groups = "drop") %>%
  mutate(db = "clustrepdb")

size_df <- tibble(
  db = c("nr", "clustnr", "repdb", "clustrepdb"),
  n = c(nr_totals[["nr"]], nr_totals[["clustnr"]], N$n_seqs, N$n_clust)
)

reduction_repdb <- repdb_df %>%
  group_by(c) %>%
  summarise(num_seqs = sum(num_seqs, na.rm = TRUE), db = "repdb", .groups = "drop") %>%
  bind_rows(select(clust_stats, c, num_seqs, db)) %>%
  pivot_wider(names_from = db, values_from = num_seqs) %>%
  mutate(db = "repdb", size = clust_repdb, reduction = repdb - clust_repdb,
         prop_reduction = reduction / repdb) %>%
  arrange(desc(prop_reduction)) %>%
  left_join(distinct(tax, k, p, c), by = "c") %>%
  select(-repdb, -clust_repdb)
```

## Database sizes (Fig. 4A)

```{r n-db-sizes, eval=HAVE_NR}
N$n_nr      <- size_df$n[size_df$db == "nr"]
N$n_clustnr <- size_df$n[size_df$db == "clustnr"]
N$repdb_vs_nr_frac      <- N$n_seqs / N$n_nr
N$nr_clust_reduction    <- (N$n_nr - N$n_clustnr) / N$n_nr * 100
N$repdb_clust_reduction <- (N$n_seqs - N$n_clust) / N$n_seqs * 100

show_tbl(
  size_df %>%
    transmute(DB = db, Sequences = n, `Sequences (B)` = n / 1e9,
              `% of nr` = n / N$n_nr * 100),
  "Total number of sequences per database (Fig. 4A)."
)
```

```{r n-db-sizes-text, eval=HAVE_NR, echo=FALSE, results="asis"}
cat(sprintf(
  "RepDB is substantially more compact than nr, including **%s as many sequences (%s versus %s)**. Clustering removes %s of nr but only %s of RepDB, and the two clustered databases end up at %s (ClusteredNR) and %s (ClusteredRepDB).\n",
  paste0(fmt_d(1 / N$repdb_vs_nr_frac, 2), "x fewer"),
  fmt_b(N$n_seqs), fmt_b(N$n_nr),
  fmt_p(N$nr_clust_reduction), fmt_p(N$repdb_clust_reduction),
  fmt_b(N$n_clustnr), fmt_b(N$n_clust)
))
```

```{r n-ecoli, eval=HAVE_ECOLI}
ecoli_tbl <- tibble(
  DB = c("nr", "clustnr"),
  `E. coli sequences` = c(ecoli_counts[["nr"]], ecoli_counts[["clustnr"]])
) %>%
  mutate(`% of DB` = `E. coli sequences` / c(N$n_nr, N$n_clustnr) * 100)

N$ecoli_nr        <- ecoli_tbl$`E. coli sequences`[1]
N$ecoli_clustnr   <- ecoli_tbl$`E. coli sequences`[2]
N$ecoli_nr_prop   <- ecoli_tbl$`% of DB`[1]
N$ecoli_reduction <- (N$ecoli_nr - N$ecoli_clustnr) / N$ecoli_nr * 100

show_tbl(ecoli_tbl,
  "The *E. coli* problem in nr: sequences assigned to any taxid under *Escherichia coli* (562).")
```

## Taxonomic evenness (Fig. 4B)

```{r n-gini, eval=HAVE_NR}
gini_df <- bind_rows(nr_classes, clustnr_classes, repdb_classes, clustrepdb_classes) %>%
  group_by(k, db) %>%
  summarise(gini_index = gini(n), .groups = "drop")

show_tbl(
  gini_df %>%
    pivot_wider(names_from = db, values_from = gini_index) %>%
    rename(Kingdom = k) %>%
    select(Kingdom, any_of(names(color_benchdbs))),
  "Gini index of the class-level sequence distribution per kingdom and database (0 = perfectly even, 1 = maximally skewed).",
  digits = 3
)
```

## Where the redundancy is removed (Fig. 4C)

```{r reduction-rank, eval=HAVE_NR}
rank_reduction <- function(df) {
  df %>%
    filter(!is.na(reduction)) %>%
    arrange(desc(reduction)) %>%
    mutate(Rank = row_number(),
           cum_reduction = cumsum(reduction),
           cum_share = cum_reduction / sum(reduction, na.rm = TRUE) * 100)
}
rank_nr    <- rank_reduction(reduction_nr)
rank_repdb <- rank_reduction(reduction_repdb)

N$nr_top4_classes    <- paste(head(rank_nr$c, 4), collapse = ", ")
N$nr_top4_share      <- rank_nr$cum_share[4]
N$repdb_top4_classes <- paste(head(rank_repdb$c, 4), collapse = ", ")
N$repdb_top4_share   <- rank_repdb$cum_share[4]
N$repdb_top_prop_classes <- paste(head(arrange(reduction_repdb, desc(prop_reduction))$c, 3),
                                  collapse = ", ")
rank_of <- function(df, cls) {
  i <- which(df$c == cls); if (length(i)) df$Rank[i[1]] else NA_integer_
}
N$rank_bacilli_repdb     <- rank_of(rank_repdb, "Bacilli")
N$rank_bacteroidia_repdb <- rank_of(rank_repdb, "Bacteroidia")
```

```{r n-reduction-nr, eval=HAVE_NR}
show_tbl(
  rank_nr %>% head(15) %>%
    transmute(Rank, Kingdom = k, Class = c, Removed = reduction,
              `% reduced` = prop_reduction * 100,
              `Cumulative % of all nr reduction` = cum_share),
  "Classes contributing most to the redundancy clustering removes from nr (top 15)."
)
```

```{r n-reduction-repdb-abs, eval=HAVE_NR}
show_tbl(
  rank_repdb %>% head(15) %>%
    transmute(Rank, Kingdom = k, Phylum = p, Class = c, Removed = reduction,
              `% reduced` = prop_reduction * 100,
              `Cumulative % of all RepDB reduction` = cum_share),
  "Classes contributing most to the redundancy clustering removes from RepDB (top 15)."
)
```


```{r n-reduction-repdb-rel, eval=HAVE_NR}
show_tbl(
  reduction_repdb %>%
    arrange(desc(prop_reduction)) %>% head(15) %>%
    transmute(Kingdom = k, Phylum = p, Class = c,
              `Clustered size` = size, Removed = reduction,
              `% reduced` = prop_reduction * 100),
  "Most redundant RepDB classes in relative terms (top 15 by proportion of sequences removed)."
)
```

The most significant reductions in relative terms occur in
**`r if (HAVE_NR) N$repdb_top_prop_classes else "-"`**.

```{r n-caudo-nr, eval=HAVE_NR}
show_tbl(
  nr_classes %>%
    filter(k == "Viruses") %>%
    mutate(Group = if_else(c == "Caudoviricetes", "Caudoviricetes", "other viruses")) %>%
    group_by(Group) %>%
    summarise(Sequences = sum(n), .groups = "drop") %>%
    mutate(`% sequences` = Sequences / sum(Sequences) * 100),
  "Caudoviricetes share of the viral sequences in nr (compare with the RepDB viral table above)."
)
```

## Fig. 4

```{r fig4, eval=HAVE_NR, fig.width=9, fig.height=3}
plot_gini <- gini_df %>%
  mutate(db = factor(db, levels = names(color_benchdbs))) %>%
  ggplot(aes(db, gini_index, fill = k)) +
  geom_line(aes(group = k), color = "grey40", linetype = 2) +
  geom_point(pch = 21, size = 3) +
  scale_fill_manual(values = color_kingdoms) +
  scale_y_continuous(limits = c(0.65, 1)) +
  guides(fill = guide_legend(nrow = 2, byrow = TRUE, override.aes = list(size = 2))) +
  theme(
    legend.position = c(0.32, 0.15),
    legend.key.spacing.y = unit(0, "mm"),
    legend.key.spacing.x = unit(0, "mm"),
    legend.key.size = unit(3, "mm"),
    legend.key = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    legend.text = element_text(size = 6, margin = unit(c(0, 0, 0, .25), "mm")),
    legend.title = element_text(size = 7, margin = unit(c(0, 0, .5, 0), "mm"))
  ) +
  labs(x = "", y = "Gini Index", fill = "Kingdom")

plot_reduction <- bind_rows(reduction_nr, reduction_repdb) %>%
  ggplot(aes(prop_reduction, k)) +
  ggbeeswarm::geom_quasirandom(pch = 21, aes(fill = k, size = size),
                               stroke = .25, alpha = .75) +
  ggrepel::geom_text_repel(
    data = reduction_repdb %>% group_by(k) %>% top_n(3, prop_reduction),
    aes(label = c), size = 1.5,
    segment.curvature = -0.1, segment.ncp = 3, segment.angle = 10
  ) +
  facet_wrap(~db, scales = "free_x") +
  labs(x = "% Reduced", y = "") +
  scale_size(range = c(.5, 5)) +
  scale_fill_manual(values = color_kingdoms) +
  theme(legend.position = "none")

plot_size <- size_df %>%
  mutate(clustered = grepl("clust", db), db = gsub("clust", "", db)) %>%
  ggplot(aes(db, n, fill = db, alpha = clustered)) +
  geom_col(position = "dodge") +
  scale_alpha_manual(values = c(1, 0.5)) +
  scale_fill_manual(values = color_benchdbs) +
  scale_y_continuous(expand = c(0, 0.05),
                     labels = scales::unit_format(unit = "B", scale = 1e-9)) +
  labs(x = "", y = "N") +
  theme(legend.position = "none")

plot_red <- plot_size + plot_gini + plot_reduction +
  plot_layout(widths = c(1, 3.7, 7)) +
  plot_annotation(tag_levels = "A")

plot_red
```

# Exploratory

```{r euka-cooc}
show_tbl(
  euka_cooc %>%
    arrange(desc(n)) %>%
    transmute(`Proteome A` = q, `Phylum A` = p.x,
              `Proteome B` = t, `Phylum B` = p.y, `Co-occurrence` = n) %>%
    head(20),
  sprintf("Eukaryote-eukaryote co-occurrence across phyla (top 20 of %s pairs above the step-1 threshold of %s).",
          fmt_n(nrow(euka_cooc)), fmt_n(bundle_meta$params$euka_cooc_min))
)
```

```{r viral-cooc}
show_tbl(head(viral_cooc, 20), "Viral classes co-occurring with non-viral clades (top 20).")
```

```{r viral-cooc-megavir}
show_tbl(head(viral_cooc_megavir, 20), "Megaviricetes co-occurring with Diaphoretickes (top 20).")
```


# Figures written to disk

```{r comparison-panel}
have_rds <- function(p) !is.null(p) && nzchar(p) && file.exists(.rp(p))
if (have_rds(P$comparison_rds)) {
  plot_cohen <- readRDS(.rp(P$comparison_rds)) + plot_annotation(tag_levels = "A")
} else {
  message("comparison.Rds not found; run comparison.Rmd first to include Fig. 5")
}
if (have_rds(P$recall_rds)) {
  recall_plot <- readRDS(.rp(P$recall_rds))
} else {
  message("recall.RDS not found; Fig. S1 will not be written")
}
```

```{r save-fig-paper}
save_if <- function(name, obj, ...) {
  if (exists(obj, inherits = TRUE)) {
    ggsave(fig(name), get(obj, inherits = TRUE), ...)
    tibble(Figure = name, Status = "written")
  } else {
    tibble(Figure = name, Status = sprintf("skipped (`%s` not available)", obj))
  }
}

show_tbl(
  bind_rows(
    save_if("Fig2.pdf", "p4", width = 11, height = 11),
    save_if("Fig3.pdf", "plot_cont", width = 9, height = 6),
    save_if("Fig4.pdf", "plot_red", height = 3, width = 9),
    # save_if("Fig5.pdf", "plot_cohen", width = 6.5, height = 4.5),
    # save_if("FigS1.pdf", "recall_plot", width = 4, height = 4),
    save_if("FigS2.pdf", "pvir2", width = 4.3, height = 4.5)
  ),
  sprintf("Figures written to `%s`.", figdir)
)
```

# Session info

```{r session-info}
sessionInfo()
```
