---
title: "Exploring the available proteomes"
date: "`r format(Sys.time(), '%d-%b-%Y')`"
output:
    rmdformats::html_clean:
        number_sections: false
        code_folding: show
        self_contained: true
        toc_depth: 3
        highlight: kate
params:
    available: "results/meta/available_proteomes.tsv"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, warning = FALSE, message = FALSE)
```

This is a **boilerplate** notebook for exploring `available_proteomes.tsv`, the
menu of everything RepDBmaker could include, produced by:

```bash
snakemake -j1 --until available_proteomes
```

Use it to understand what is available before choosing a subset, and to write an
`ids` file for a custom database. Adapt the filters to your question — the last
section shows how to export a selection. Point `params$available` at your file
(default `results/meta/available_proteomes.tsv`); knit from the repository root.

```{r libs}
suppressPackageStartupMessages(library(tidyverse))
theme_set(theme_classic())

# shared helpers (rp, ...) and palettes (color_sources); knit from the repo root
source("workflow/scripts/functions.R")
source("workflow/scripts/palettes.R")
```

# Load and annotate

The table has one row per proteome: `mnemo`, `Completeness`, `Data_type`, and the
seven taxonomic ranks `k, p, c, o, f, g, s` (domain → species). We add a
`source` (from the mnemo prefix and domain) and a numeric completeness.

```{r load}
prot <- read_tsv(rp(params$available), col_types = cols(.default = "c")) %>%
    mutate(
        Completeness = suppressWarnings(as.numeric(Completeness)),
        domain = na_if(k, ""),
        source = case_when(
            str_starts(mnemo, "UP")  ~ "uniprot",
            str_starts(mnemo, "EP")  ~ "eukprot",
            str_starts(mnemo, "P10") ~ "p10k",
            str_starts(mnemo, "CUS") ~ "custom",
            k == "Viruses"           ~ "virus",
            k %in% c("Bacteria", "Archaea") ~ "gtdb",
            TRUE ~ "other"
        )
    )

glimpse(prot)
```

# Overview: how many, from where

```{r overview-table}
prot %>%
    count(domain, source, name = "n") %>%
    arrange(desc(n)) %>%
    knitr::kable(caption = "Proteomes per domain and source")
```

```{r overview-plot, fig.width=8, fig.height=4}
prot %>%
    count(source, domain) %>%
    ggplot(aes(n, fct_reorder(domain, n, sum), fill = source)) +
    geom_col() +
    scale_fill_manual(values = color_sources) +
    scale_x_continuous(labels = scales::label_number(big.mark = ",")) +
    # scale_x_log10() +
    labs(x = "proteomes", y = NULL, title = "Available proteomes by source and domain")
```

# Data types

Different sources contribute different kinds of data (genomes, transcriptomes,
single-cell, …). This matters when interpreting completeness and coverage.

```{r data-type, fig.width=8, fig.height=4}
prot %>%
    count(source, Data_type) %>%
    ggplot(aes(n, fct_reorder(Data_type, n, sum), fill = source)) +
    geom_col() +
    scale_fill_manual(values = color_sources) +
    scale_x_continuous(labels = scales::label_number(big.mark = ",")) +
    labs(x = "proteomes", y = NULL, title = "Data types per source")
```

# Taxonomic composition

Top phyla per domain (edit `top_n` / the rank you group by). Because the sources
use different frameworks, "phylum" (`p`) is not the same concept across domains
(GTDB vs UniEuk vs ICTV) — see the harmonization QC report under `results/qc/`.

```{r taxonomy, fig.width=9, fig.height=6}
prot %>%
    filter(!is.na(domain), p != "") %>%
    count(domain, p) %>%
    group_by(domain) %>%
    slice_max(n, n = 12) %>%
    ungroup() %>%
    ggplot(aes(n, reorder(p, n), fill = domain)) +
    geom_col(show.legend = FALSE) +
    scale_x_continuous(labels = scales::label_number(big.mark = ",")) +
    facet_wrap(~domain, scales = "free") +
    labs(x = "proteomes", y = NULL, title = "Top phyla per domain")
```

# Completeness

Completeness is numeric where available (CheckM2 for prokaryotes, BUSCO for
eukaryotes) and missing for viruses. Remember BUSCO scores are **not comparable
across lineages** (lineage-specific ortholog sets), so compare within a group.

```{r completeness, fig.width=8, fig.height=4}
prot %>%
    filter(!is.na(Completeness)) %>%
    ggplot(aes(Completeness, fct_reorder(source, Completeness), fill = source)) +
    geom_boxplot(show.legend = FALSE, outlier.size = 0.5) +
    scale_fill_manual(values = color_sources) +
    labs(x = "completeness", y = NULL, title = "Completeness distribution by source")
```

# Subsetting: build an `ids` file

Boilerplate for turning a selection into an `ids` file usable as a custom
database (`dbs.build.custom.<name>.ids` in the config). Edit the `filter()` to
your question.

```{r subset}
selection <- prot %>%
    filter(
        domain == "Eukaryota",
        # example: keep reasonably complete proteomes only
        is.na(Completeness) | Completeness >= 70,
        # example: restrict to a clade (edit rank/value)
        # p == "Discoba"
    )

nrow(selection)
count(selection, source)
```

```{r write-ids, eval=FALSE}
# write one mnemo per line; point the config's `ids:` at this file
selection %>%
    distinct(mnemo) %>%
    write_tsv("resources/my_selection.ids", col_names = FALSE)
```

# Session info

```{r session}
sessionInfo()
```
